\documentclass[10pt,twoside,openright]{book}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amssymb}
\usepackage{amsthm}
\usepackage{mathpazo}          % Palatino text + matching math
\usepackage[scaled=0.85]{helvet}  % Helvetica for sans-serif (headers, boxes)
\usepackage{courier}           % Courier for monospace (code listings)
\linespread{1.05}              % Palatino reads best with slightly opened leading
% 6×9 inch trim — standard academic textbook
% Inner (gutter) margin wider to account for binding
\usepackage[paperwidth=6in,paperheight=9in,
            top=0.875in,bottom=0.9in,
            inner=1.0in,outer=0.75in,
            headsep=0.25in,footskip=0.4in]{geometry}
\usepackage{fancyhdr}   % running headers
\usepackage{microtype}
% Traditional book paragraphs: indented first line, no inter-paragraph space
\setlength{\parindent}{1.5em}
\setlength{\parskip}{0pt plus 1pt}
\usepackage{enumitem}
\usepackage{array}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage[font=small,labelfont=bf]{caption}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage[most]{tcolorbox}
\usepackage{listings}
\usepackage[round,authoryear,sort&compress]{natbib}
\PassOptionsToPackage{hyphens}{url}   % allow long URLs to break at hyphens
\usepackage[hidelinks,plainpages=false,pdfpagelabels]{hyperref}

\usetikzlibrary{arrows.meta,backgrounds,decorations.pathmorphing,fit,matrix,patterns,positioning,shapes.geometric}

% ── Chapter and section heading styles ───────────────────────────────────
\usepackage{titlesec}

\titleformat{\chapter}[display]
  {\normalfont\Large\bfseries\sffamily}        % format
  {\chaptertitlename\ \thechapter}              % label
  {12pt}                                        % sep between label and title
  {\LARGE}                                      % before-code (title font)

\titleformat{\section}
  {\normalfont\large\bfseries\sffamily}
  {\thesection}{1em}{}

\titleformat{\subsection}
  {\normalfont\normalsize\bfseries\sffamily}
  {\thesubsection}{1em}{}

\titlespacing*{\chapter}{0pt}{30pt}{20pt}
\titlespacing*{\section}{0pt}{18pt plus 4pt minus 2pt}{6pt plus 2pt}
\titlespacing*{\subsection}{0pt}{12pt plus 3pt minus 1pt}{4pt plus 1pt}

% ── Part page styling ────────────────────────────────────────────────────
\titleformat{\part}[display]
  {\centering\normalfont\Huge\bfseries\sffamily}
  {\partname\ \thepart}
  {12pt}
  {\Huge}

% ── Cover colour palette ──────────────────────────────────────────────────
\definecolor{cvBG}{HTML}{F8FBFF}
\definecolor{cvNavy}{HTML}{0D1B2A}
\definecolor{cvDarkBlue}{HTML}{023E8A}
\definecolor{cvMidBlue}{HTML}{0353A4}
\definecolor{cvOcean}{HTML}{0077B6}
\definecolor{cvTeal}{HTML}{00B4D8}
\definecolor{cvTealLight}{HTML}{48CAE4}
\definecolor{cvFieldBG}{HTML}{E8F4FD}
\definecolor{cvRule}{HTML}{CBD5E0}
\definecolor{cvSubtitle}{HTML}{2D5986}
\definecolor{cvBoundary}{HTML}{0096C7}

\title{Machine Learning by Design: From Problem Framing to Reliable Systems}
\author{Sam Urmian}
\date{}

\hypersetup{
  pdftitle={Machine Learning by Design: From Problem Framing to Reliable Systems},
  pdfauthor={Sam Urmian},
  pdfsubject={Machine Learning, Evaluation, and Reliable AI Systems}
}

\captionsetup{
  format=plain,
  justification=raggedright,
  singlelinecheck=false
}
\setlength{\abovecaptionskip}{6pt}
\setlength{\belowcaptionskip}{2pt}
\setlength{\textfloatsep}{16pt plus 2pt minus 3pt}
\setlength{\floatsep}{12pt plus 2pt minus 2pt}
\setlength{\intextsep}{12pt plus 2pt minus 2pt}
\renewcommand{\arraystretch}{1.08}
\newcolumntype{L}[1]{>{\raggedright\arraybackslash}p{#1}}

\definecolor{codebg}{RGB}{247,247,247}
\definecolor{codecomment}{RGB}{90,90,90}
\definecolor{codekeyword}{RGB}{0,92,175}
\definecolor{codestring}{RGB}{163,21,21}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[chapter]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[chapter]
\newtheorem{example}{Example}[chapter]

\lstdefinestyle{bookpython}{
  language=Python,
  basicstyle=\ttfamily\small,
  keywordstyle=\color{codekeyword}\bfseries,
  commentstyle=\color{codecomment}\itshape,
  stringstyle=\color{codestring},
  showstringspaces=false,
  breaklines=true,
  frame=single,
  backgroundcolor=\color{codebg},
  columns=fullflexible
}

\lstnewenvironment{codeexample}[1][]
  {\lstset{style=bookpython,#1}}
  {}

\newenvironment{learninggoals}
  {\section*{Learning Goals}\begin{itemize}[leftmargin=*]}
  {\end{itemize}}

\newenvironment{studyquestions}
  {\section*{Study Questions}\begin{enumerate}[leftmargin=*,label=\arabic*.]}
  {\end{enumerate}}

\newenvironment{chaptersummary}
  {\section*{Chapter Summary}\begin{itemize}[leftmargin=*]}
  {\end{itemize}}

\newenvironment{exercises}
  {\section*{Exercises}\begin{enumerate}[leftmargin=*,label=\arabic*.]}
  {\end{enumerate}}

% ── Box colour palette (four families) ────────────────────────────────────
% Blue  — conceptual / analytical
\definecolor{boxBlue}{HTML}{1565C0}
\definecolor{boxBlueBG}{HTML}{E8F0FE}
% Amber — warnings / failure modes
\definecolor{boxAmber}{HTML}{E65100}
\definecolor{boxAmberBG}{HTML}{FFF3E0}
% Green — practical / decisions
\definecolor{boxGreen}{HTML}{2E7D32}
\definecolor{boxGreenBG}{HTML}{E8F5E9}
% Gray  — advanced / supplementary
\definecolor{boxGray}{HTML}{546E7A}
\definecolor{boxGrayBG}{HTML}{ECEFF1}

% ── Base box style ────────────────────────────────────────────────────────
\tcbset{
  mlbox/.style={
    enhanced,
    breakable,
    sharp corners,
    boxrule=0pt,
    frame hidden,
    borderline west={2.5pt}{0pt}{#1},
    colback=#1BG,
    left=8pt,
    right=8pt,
    top=6pt,
    bottom=6pt,
    before skip=10pt plus 2pt,
    after skip=10pt plus 2pt,
    fontupper=\small,
    parbox=false,
  },
  mlbox/.default={boxBlue}
}

% ── Blue family (conceptual / analytical) ─────────────────────────────────
\newtcolorbox{mathlens}{mlbox=boxBlue, before upper={\textbf{Math Lens.}\ }}
\newtcolorbox{thinkingtool}{mlbox=boxBlue, before upper={\textbf{Problem-Solving Technique.}\ }}
\newtcolorbox{problemsolvinglens}{mlbox=boxBlue, before upper={\textbf{Problem-Solving Lens.}\ }}
\newtcolorbox{complexitylens}{mlbox=boxBlue, before upper={\textbf{Complexity Lens.}\ }}
\newtcolorbox{sanitycheck}{mlbox=boxBlue, before upper={\textbf{Sanity Check.}\ }}
\newtcolorbox{claimdefense}{mlbox=boxBlue, before upper={\textbf{Claim Defense.}\ }}
\newtcolorbox{whychaptermatters}{mlbox=boxBlue, before upper={\textbf{Why This Part Matters.}\ }}

% ── Amber family (warnings / failures) ────────────────────────────────────
\newtcolorbox{warningbox}{mlbox=boxAmber, before upper={\textbf{Warning.}\ }}
\newtcolorbox{failurebox}{mlbox=boxAmber, before upper={\textbf{Failure Box.}\ }}
\newtcolorbox{failurepattern}{mlbox=boxAmber, before upper={\textbf{Failure Pattern.}\ }}

\newenvironment{mistakebox}
  {\begin{tcolorbox}[mlbox=boxAmber, before upper={\textbf{Mistake Checklist.}}]\begin{itemize}[leftmargin=*]}
  {\end{itemize}\end{tcolorbox}}

% ── Green family (practical / decisions) ──────────────────────────────────
\newtcolorbox{practicalnote}{mlbox=boxGreen, before upper={\textbf{Practical Note.}\ }}
\newtcolorbox{decisionbox}{mlbox=boxGreen, before upper={\textbf{Decision Box.}\ }}
\newtcolorbox{evidencebox}{mlbox=boxGreen, before upper={\textbf{Evidence Box.}\ }}
\newtcolorbox{prereqbridge}{mlbox=boxGreen, before upper={\textbf{Prerequisite Bridge.}\ }}

% ── Gray family (advanced / supplementary) ────────────────────────────────
\newtcolorbox{advancednote}{mlbox=boxGray, before upper={\textbf{Advanced Note.}\ }}

% ── Challenge exercises (unchanged) ───────────────────────────────────────
\newenvironment{challengeexercises}
  {\section*{Challenge Exercises}\begin{enumerate}[leftmargin=*,label=\arabic*.]}
  {\end{enumerate}}

% ── Running headers (twoside) ─────────────────────────────────────────────
\pagestyle{fancy}
\fancyhf{}
\fancyhead[LE]{\small\itshape Machine Learning by Design}
\fancyhead[RO]{\small\itshape\nouppercase{\leftmark}}
\fancyfoot[LE,RO]{\small\thepage}
\fancypagestyle{plain}{%
  \fancyhf{}
  \fancyfoot[LE,RO]{\small\thepage}
  \renewcommand{\headrulewidth}{0pt}
}
\renewcommand{\headrulewidth}{0.4pt}

% ── Typography tolerance ─────────────────────────────────────────────────
\tolerance=1200
\hbadness=1200
\emergencystretch=3em

\begin{document}

\hypersetup{pageanchor=false}

% ── Cover (full-bleed, no page number) ────────────────────────────────────
\input{frontmatter/cover2}

\frontmatter
\input{frontmatter/title}
\cleardoublepage
\hypersetup{pageanchor=true}
\input{frontmatter/preface}
\tableofcontents

\mainmatter
\part*{Mathematical Bridge: Just-in-Time Review}
\input{chapters/bridge01-linear-algebra}
\input{chapters/bridge02-probability}
\part{Foundations}
\input{chapters/ch01}
\input{chapters/ch02}
\input{chapters/ch02b}
\input{chapters/ch03}
\input{chapters/ch04}
\input{chapters/ch04b}
\input{chapters/part-synthesis-foundations}
\part{Models and Representations}
\input{chapters/ch05}
\input{chapters/ch05b}
\input{chapters/ch06}
\input{chapters/ch06a}
\input{chapters/ch06b}
\input{chapters/part-synthesis-models-representations}
\part{Applications and Modalities}
\input{chapters/ch07}
\input{chapters/ch08}
\input{chapters/ch09}
\input{chapters/ch09a}
\input{chapters/part-synthesis-applications-modalities}
\part{Evidence, Reliability, and Systems}
\input{chapters/ch10}
\input{chapters/ch11}
\input{chapters/ch12}
\input{chapters/part-synthesis-evidence-systems}

\appendix
\input{appendices/math-foundations}

\backmatter
\input{backmatter/conclusion}
\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
