\chapter*{Preface}
\addcontentsline{toc}{chapter}{Preface}

Modern AI is often taught in two unsatisfying ways. One approach presents the machine-learning core of the field as a catalog of methods: one chapter on linear models, another on trees, another on neural networks, another on transformers. Another turns it into a toolkit course: load a library, fit a model, report a score, move on. Both approaches miss the deeper intellectual habit. Real model-based AI work is not mainly about collecting algorithms. It is about solving problems under uncertainty, with incomplete data, imperfect proxies, constrained decisions, and evidence that must be defended.

This book is written around that stronger idea. It does not ask first, ``Which model is famous?'' It asks, ``What is the problem, what evidence do we have, what claim are we making, and what would justify trusting the result?'' The book is therefore problem-driven rather than algorithm-driven. Methods matter, but they appear as answers to modeling dilemmas, not as museum exhibits.

This book is for undergraduate students, teachers, and serious self-learners who want a coherent first serious introduction to machine learning, evaluation, and reliable AI systems. It assumes curiosity, some programming experience, and enough mathematical maturity to follow equations when they are explained carefully. The volume includes two mathematical bridge chapters on linear algebra and on probability, risk, and learning signals, but most first-course readers should begin with Chapter 1 and then use the bridge chapters as just-in-time review whenever notation or derivations start to feel thin. The main line is undergraduate-first. Readers who want a faster or deeper route can go further through prerequisite bridges, advanced notes, challenge exercises, and companion code.

That scope matters. This is not a full survey of artificial intelligence in the broad classical sense. It is first a machine-learning textbook with an AI-systems endgame. Topics such as search, planning, and symbolic reasoning appear only for orientation when they clarify where machine learning fits. Reinforcement learning receives a full chapter because the feedback-loop and exploration-exploitation logic it introduces connects directly to recommendation, deployment, and system-level reasoning later in the book. The book's center of gravity is the machine-learning core of modern AI---including probabilistic models, generative models, and recommendation systems---together with the habits needed to evaluate, stress-test, and operationalize model-based systems.

The title is broad because the problems are broad, but the shelf claim is precise. This book should be read as a machine-learning text about framing, evidence, reliability, and systems, not as a general survey of all of AI.

\section*{Suggested Reading Paths}

\begin{itemize}[leftmargin=*]
\item \textbf{First machine-learning course.} Read Chapter 1 first. Then continue in order, using the Mathematical Bridge chapters only when later chapters point back to them or when the notation starts to slow you down.
\item \textbf{Math-refresh path.} Read Chapter 1 first for the problem-framing spine, then work through the two Mathematical Bridge chapters before continuing to Chapter 2.
\item \textbf{One-semester teaching path.} Chapters~\ref{ch:framing-learning-problems}--\ref{ch:dataset-design} form the core. Add Chapter~\ref{ch:optimization-neural-networks} on optimization and neural networks. If probabilistic foundations are a priority, include Chapter~\ref{ch:probabilistic-models} on probabilistic models and uncertainty. Then add Chapter~\ref{ch:representation-foundation-models} on representation learning and foundation models. Choose the modality chapter most relevant to the course (Chapters~\ref{ch:vision}--\ref{ch:audio-time}), and if recommendation or ranking is relevant, include Chapter~\ref{ch:ranking-recommendation}. Finish with Chapters~\ref{ch:experiments-claims}--\ref{ch:models-systems} for experiments, reliability, and systems.
\item \textbf{Probabilistic and generative track.} For a course emphasizing probabilistic reasoning and generative models, the key arc is: Bridge~II (probability), Chapter~\ref{ch:structure-beyond-linearity} (Naive Bayes introduction), Chapter~\ref{ch:probabilistic-models} (GMMs, EM, Bayesian reasoning, uncertainty), Chapter~\ref{ch:representation-foundation-models} (autoencoders and learned representations), Chapter~\ref{ch:generative-representations} (VAEs, GANs, diffusion models), and then forward to Chapter~\ref{ch:reliability} (uncertainty quantification for reliability).
\end{itemize}

The organizing framework of the book is simple and reusable. Every serious AI project eventually has to answer seven questions:
\begin{enumerate}[leftmargin=*,label=\arabic*.]
\item What decision are we actually supporting?
\item What learning task is the closest defensible proxy for that decision?
\item What data stands in for the world, and what could bias or leak it?
\item What representation makes the relevant structure visible?
\item What is the simplest serious baseline?
\item What evidence would justify the claim?
\item How will the model output become an action inside a real system?
\end{enumerate}

Chapters 1 through 3 introduce the technical core of that method through task framing, predictive judgment, and evaluation discipline. Later chapters extend the same logic outward. Experiments become arguments. Thresholds become policies. Reliability becomes a property of the whole pipeline. Deployment becomes part of the modeling problem rather than a final packaging step. The final part of the book is therefore not an ethics appendix. It is where evidence, reliability, and systems work are brought into the same technical frame.

The book also returns repeatedly to one institutional setting: course support. The early chapters use it for framing, metric choice, and baseline design. The later language, experiments, reliability, and systems chapters turn that same setting into a grounded support assistant with retrieval, escalation, and deployment constraints. The back half also cross-checks those same habits against a dashboard-camera pedestrian detector first developed in the vision chapter, so the systems endgame does not read as if it belongs only to educational NLP.

To make those habits teachable, the book uses a small set of recurring problem-solving techniques. Some are cross-cutting across the whole book, while others become especially important in modalities such as vision, language, sequential prediction, or transfer learning.
\begin{itemize}[leftmargin=*]
\item \textbf{Decision-first framing}: start from the action and constraint, not from the model.
\item \textbf{Proxy audit}: separate the real goal from the measurable target.
\item \textbf{Representation audit}: ask what information the representation must preserve before blaming the model family.
\item \textbf{Baseline ladder}: begin with the simplest serious baseline, then justify each increase in complexity.
\item \textbf{Claim audit}: state the exact claim before interpreting the experiment.
\item \textbf{Slice diagnosis}: inspect where performance fails before trusting any aggregate metric.
\item \textbf{Threshold as policy}: treat thresholds, abstention, and review paths as operational choices.
\item \textbf{Pipeline realism}: ask what breaks between notebook success and system success.
\item \textbf{Reuse audit}: separate gains from pretrained representation, downstream head, retrieval, prompting, or fine-tuning.
\item \textbf{Invariance audit}: ask what kinds of variation should and should not change the answer.
\item \textbf{Context audit}: ask how much context and grounding the task really requires.
\item \textbf{Prediction-time audit}: ask what information is truly available at decision time and what future information must stay forbidden.
\end{itemize}

The book uses a small set of recurring teaching devices:
\begin{itemize}[leftmargin=*]
\item \textbf{Problem-Solving Lens}: the short framing question that tells the reader what kind of modeling decision the chapter is really about.
\item \textbf{Problem-Solving Technique}: a reusable reasoning move for choosing, debugging, or redesigning a method.
\item \textbf{Math Lens}: one compact derivation, theorem, or identity that explains why the method behaves as it does.
\item \textbf{From Scratch}: a tiny implementation, usually in plain Python, that exposes the mechanism without relying on a library.
\item \textbf{Failure Pattern}: one common way a seemingly reasonable method can mislead when the representation, metric, or claim is weak.
\item \textbf{Chapter Artifact}: one memo, card, sheet, or brief that turns the chapter's reasoning into a reusable deliverable.
\end{itemize}
The companion labs carry the larger implementations. The book uses small executable fragments and short proofs so that the mechanism becomes visible before software details take over. Each chapter should also leave the reader with a reusable artifact such as a memo, claim sheet, review card, or readiness brief.

These are not slogans. They are the actual habits that connect introductory machine learning to serious practice, advanced coursework, empirical research, and deployment work. A student who learns only formulas and architectures may still fail badly on real problems. A student who learns how to frame, test, diagnose, and redesign an AI system can carry that method across domains.

The book is not trying to replace the classical references that already define much of the field's technical backbone. Texts such as \citet{bishop2006}, \citet{hastie2009}, \citet{goodfellow2016}, and \citet{murphy2022} remain essential. What this book changes is the teaching order, the recurring problem-solving language, and the explicit bridge from model construction to evidence and system design.

The book is paired with companion code, but the two are designed to age differently. The book emphasizes concepts, notation, pseudocode, and durable examples. The repository carries the runnable implementations. When libraries evolve, the code can be updated without forcing the text to depend on transient APIs or fashionable tooling.

The book is organized as an unnumbered mathematical bridge followed by four numbered parts: foundations, models and representations, applications and modalities, and evidence, reliability, and systems. That structure is deliberate. The bridge chapters make the notation and training language legible without turning the book into a theorem-first text. The foundations build the problem-solving core: framing, prediction quality, decision rules, evaluation discipline, baselines, and early model families including data design. The middle parts move from optimization and neural networks through probabilistic models, representation learning, and generative models to modality-specific applications (vision, language, sequential data) and recommendation systems. The final chapters widen the unit of analysis from models to experiments, reliability, and systems.

Each chapter follows a common rhythm: a modeling dilemma, a problem-solving lens, worked examples, one or more of the recurring teaching devices above, common failure modes, and exercises designed to strengthen four habits of mind: checking concepts, building small solutions, diagnosing failure, and designing better systems. The goal is not to make AI easy by removing rigor. The goal is to make rigor usable.

\section*{About This Project}

This project started as part of NOKI, the AI Olympiad in Norway, to train the Norwegian national team for the International Olympiad in Artificial Intelligence (IOAI). NOKI is directed by Sam Urmian, the author of this book, who led the programme during the period in which this manuscript was developed. Sam Urmian is a researcher at the University of Bergen, Norway.

The manuscript is also publicly available in the project repository at \url{https://github.com/mlgorithm/ml-by-design}.

For any questions, corrections, or feedback regarding the content of this book, please contact the author at \texttt{sam.urmian@uib.no}.
