\chapter*{Mathematical Bridge I: Linear Algebra for AI Thinking}
\addcontentsline{toc}{chapter}{Mathematical Bridge I: Linear Algebra for AI Thinking}

In AI, representation comes before model choice. This bridge gives you the vocabulary to make representation precise. After reading it, you should be able to read vectors, matrices, dot products, and shapes as modeling statements rather than isolated symbols. You should be able to compute a weighted score by hand and explain what each term contributes. You should be able to tell distance, similarity, and scale apart, and say why each matters for downstream model behavior. You should be able to interpret norms, linear maps, and projection as tools for controlling, comparing, and fitting representations. And you should be able to produce a one-page representation card before claiming that a model family fits a task.

\begin{prereqbridge}
Use this chapter as just-in-time review. Read it now if linear-algebra notation feels rusty, or come back to it when later chapters start leaning on vectors, matrices, and geometry. The chapter assumes comfort with high-school algebra---variables, substitution, evaluating expressions---but no calculus. Familiarity with Python lists or arrays is helpful but not required. Students who have already taken a linear-algebra course should still skim the chapter, because its AI-specific vocabulary (representation, coordinate, weighted evidence) recurs throughout the book.
\end{prereqbridge}

Most students first meet linear algebra as a list of operations: add vectors, multiply matrices, solve systems. That is not the most useful entry point for AI. Here, linear algebra is the language that makes representations visible. It tells us what counts as one example, what counts as one coordinate, how many examples are being processed together, what a score really is, and what notions of similarity or geometry a method is quietly assuming.

This chapter is therefore not a miniature mathematics course but a reading course in the mathematical objects that later chapters will use constantly. If you can look at a vector or matrix and ask, ``what real thing does each row, column, or coordinate mean here?'' the mathematics is already doing its job.

\begin{problemsolvinglens}
Before asking which model family to use, ask how one case is represented, what each coordinate means, what counts as similarity, and what a weighted combination of the coordinates would mean in plain language.
\end{problemsolvinglens}

\begin{thinkingtool}
\textbf{Representation audit.} Before blaming or praising a model, ask four questions: what counts as one example, what each coordinate means, what information has already been discarded, and what notion of similarity the representation makes easy.
\end{thinkingtool}

Keep the chapter artifact in view from the start. The later sections introduce shapes, weighted scores, geometry, and projection, but the payoff is a short \emph{representation card}: what one example is, what the coordinates mean, what geometry matters, and what a linear score would mean in plain language.

\section*{Opening Case: One Student, One Email, One Image Patch}

Suppose we want to support three different AI tasks:

\begin{itemize}[leftmargin=*]
\item detect which students may need early human outreach,
\item estimate whether an email is spam,
\item estimate how many animals appear in a small aerial image crop.
\end{itemize}

These tasks look different on the surface, but they share a structural step. Each one needs a way to turn one case into a mathematical object that a model can read. A student becomes a list of measurements. An email becomes word counts, sender cues, or contextual embeddings. An image crop becomes pixel intensities or learned feature maps.

That is the first linear-algebra lesson of the book. AI does not begin with magic; it begins with a decision about representation.

\begin{center}
\small
\begin{tabular}{p{2.33cm}p{7.62cm}}
\toprule
Task & One possible machine-readable description \\
\midrule
Student support & Attendance rate, quiz average, missing assignments, and recent platform activity \\
Spam filtering & Counts of important words, sender reputation cues, link patterns, and message length \\
Image counting & Pixel values or local visual features extracted from a small patch \\
\bottomrule
\end{tabular}

\medskip
\noindent\textbf{Three tasks, one shared question.} What counts as one example, and what mathematical object will stand in for it?
\end{center}

\section*{Vectors Are Organized Descriptions}

A vector is an ordered collection of coordinates. In AI, a vector is often just one represented example.

For the student-support task, we might record four normalized quantities for one student:
\[
x =
\begin{bmatrix}
0.92 \\
0.80 \\
0.10 \\
0.35
\end{bmatrix}
\]
where the coordinates mean:
\begin{itemize}[leftmargin=*]
\item \(0.92\): attendance rate,
\item \(0.80\): quiz average,
\item \(0.10\): fraction of assignments missed,
\item \(0.35\): recent platform inactivity score.
\end{itemize}

The notation matters less than the interpretation. The vector is not just ``four numbers.'' It is one student expressed through four chosen coordinates. A different representation choice would give a different vector and therefore a different modeling problem.

\begin{prereqbridge}
Read \(x_j\) as ``the \(j\)-th coordinate of the represented example.'' If the notation feels formal, translate it back into task language right away: attendance, quiz average, missed assignments, or inactivity.
\end{prereqbridge}

Two practical habits begin here. First, order matters. Once the first coordinate means attendance and the second means quiz average, we cannot silently swap them later. Second, scale matters. A coordinate measured from \(0\) to \(1000\) behaves very differently from one measured from \(0\) to \(1\). Later chapters on nearest neighbors, regularization, and optimization all depend on this simple point.

\section*{Shapes, Batches, and Why Dimension Errors Matter}

Vectors, datasets, and batches of images or sequences all have shapes. Shape is not bookkeeping. It states what counts as one case, what counts as one feature family, and what gets processed together. Misreading shape is one of the most common sources of silent bugs in AI code.

Three objects that appear across later chapters illustrate the point.

\textbf{One tabular row.} A single student described by four features is a vector $x \in \mathbb{R}^4$. Its shape is $(4,)$: four numbers, one case.

\textbf{A bag-of-words batch.} Suppose we process $32$ emails at once, each represented as a count vector over a vocabulary of $10{,}000$ words. The batch is a matrix of shape $(32, 10000)$: $32$ rows for $32$ cases, $10{,}000$ columns for $10{,}000$ features.

\textbf{A token-by-token embedding sequence.} A sentence tokenized into $T$ tokens, where each token is embedded into $d$ dimensions, forms a matrix of shape $(T, d)$. Here the rows are positions in time, not independent cases. Transposing the matrix swaps that meaning and breaks any downstream operation that assumes a row is a token position.

\begin{table}[t]
\centering
\small
\setlength{\tabcolsep}{4pt}
\begin{tabular}{p{2.72cm}p{2.38cm}p{4.85cm}}
\toprule
Object & Shape & What the dimensions mean \\
\midrule
Student feature vector & $(d,)$ & $d$ features for one student \\
Tabular dataset matrix & $(n, d)$ & $n$ examples, each with $d$ features \\
Bag-of-words batch & $(B, V)$ & $B$ documents in the batch, vocabulary size $V$ \\
Token embedding sequence & $(T, d)$ & $T$ token positions, each with a $d$-dim embedding \\
\bottomrule
\end{tabular}
\caption{Shape is a modeling statement. Rows and columns have different jobs, and confusing them causes dimension errors that are easy to miss in silent failure modes.}
\label{tab:shape-meanings}
\end{table}

A useful mini-checklist before writing any matrix operation:
\begin{itemize}[leftmargin=*]
\item What are the rows? (examples? positions? channels?)
\item What are the columns? (features? vocabulary items? time steps?)
\item What is repeated across the batch? (independent cases or steps of the same sequence?)
\item What would break if the object were transposed?
\end{itemize}

\begin{failurebox}
Dimension errors often succeed silently in Python. If a dot product or matrix multiply happens to have compatible shapes because an accidental transpose works numerically, the model trains without error messages but learns from the wrong relationship. Checking what the dimensions mean before the operation---not just that the numbers multiply---is how practitioners catch this class of bug.
\end{failurebox}

\section*{Dot Products Are Weighted Evidence}

Once an example lives in a vector, the simplest serious model asks a clean question: how much evidence does each coordinate contribute to one score?

That is the job of the dot product. If a weight vector is
\[
w =
\begin{bmatrix}
-1.6 \\
-1.1 \\
2.4 \\
1.8
\end{bmatrix},
\]
then the score \(w^\top x\) is
\[
w^\top x = (-1.6)(0.92) + (-1.1)(0.80) + (2.4)(0.10) + (1.8)(0.35).
\]
Working it out term by term gives
\[
w^\top x = -1.472 - 0.880 + 0.240 + 0.630 = -1.482.
\]

The point of this arithmetic is not the final decimal. The point is the interpretation.

\begin{itemize}[leftmargin=*]
\item Good attendance pushes the score down because its weight is negative.
\item Strong quiz performance also pushes the score down.
\item Missing assignments push the score up.
\item Inactivity also pushes the score up.
\end{itemize}

In other words, the dot product is a weighted summary of evidence.

\begin{center}
\small
\begin{tabular}{p{2.73cm}p{1.82cm}p{1.82cm}p{2.73cm}}
\toprule
Coordinate & Value & Weight & Contribution \\
\midrule
Attendance & \(0.92\) & \(-1.6\) & \(-1.472\) \\
Quiz average & \(0.80\) & \(-1.1\) & \(-0.880\) \\
Assignments missed & \(0.10\) & \(2.4\) & \(0.240\) \\
Inactivity score & \(0.35\) & \(1.8\) & \(0.630\) \\
\bottomrule
\end{tabular}

\medskip
\noindent\textbf{Reading rule.} Multiply each coordinate by how much it matters, then add the contributions.
\end{center}

\begin{mathlens}
\textbf{Why the dot product appears everywhere.} A dot product is a weighted sum, and weighted sums are the simplest way to combine many weak signals into one score. Linear regression, logistic regression, attention scores, nearest-prototype methods, and many neural-network layers all reuse this structure. The later models are richer, but the weighted-sum idea never really leaves.
\end{mathlens}

\section*{From One Example to Many: Matrices As Datasets}

Once there is more than one example, writing every case separately is wasteful. A matrix packages many vectors together.

Suppose three students are represented as rows:
\[
X =
\begin{bmatrix}
0.92 & 0.80 & 0.10 & 0.35 \\
0.70 & 0.62 & 0.30 & 0.55 \\
0.98 & 0.91 & 0.00 & 0.10
\end{bmatrix}.
\]

This matrix has shape \(3 \times 4\): three rows, four columns. In the dataset view:

\begin{itemize}[leftmargin=*]
\item rows are examples,
\item columns are features,
\item shape records how many cases and features are present.
\end{itemize}

This is one of the most useful sanity checks in practical ML. If you think a dataset has \(n\) examples and \(d\) features, then the design matrix should have \(n\) rows and \(d\) columns. Many beginner bugs are simply shape misunderstandings dressed up in fancier language.

Matrix multiplication then applies one scoring rule to every row at once:
\[
Xw =
\begin{bmatrix}
0.92 & 0.80 & 0.10 & 0.35 \\
0.70 & 0.62 & 0.30 & 0.55 \\
0.98 & 0.91 & 0.00 & 0.10
\end{bmatrix}
\begin{bmatrix}
-1.6 \\
-1.1 \\
2.4 \\
1.8
\end{bmatrix}
=
\begin{bmatrix}
-1.482 \\
-0.092 \\
-2.389
\end{bmatrix}.
\]

Now the same weighted rule has produced one score for each student.

\begin{mathlens}
If the rows of \(X\) are examples \(x_1^\top,\ldots,x_n^\top\), then
\[
Xw=
\begin{bmatrix}
x_1^\top w\\
x_2^\top w\\
\vdots\\
x_n^\top w
\end{bmatrix}.
\]
So matrix-vector multiplication is just repeated dot products, one row at a time. The matrix is not doing a different kind of math; it is packaging the same linear rule so it can be applied to many examples at once.
\end{mathlens}

\begin{codeexample}
def linear_scores(X, w, b=0.0):
    scores = []
    for row in X:
        if len(row) != len(w):
            raise ValueError("row and weight vector must have the same length")
        score = b
        for value, weight in zip(row, w):
            score += value * weight
        scores.append(score)
    return scores
\end{codeexample}

The point of this tiny function is not software convenience. It is to make the mechanism visible. A matrix is not mystical: it is a compact way to store many represented cases, and matrix multiplication applies one structured computation to every row.

\begin{example}[Matrix-Vector Product by Hand]
Consider three students, each described by two features: attendance rate and number of late submissions. The design matrix and weight vector are

\[
X =
\begin{bmatrix}
0.90 & 1 \\
0.60 & 4 \\
0.80 & 0
\end{bmatrix},
\qquad
w =
\begin{bmatrix}
-1.5 \\
2.0
\end{bmatrix}.
\]

The columns mean: first column = attendance rate, second column = number of late submissions. The weights say: good attendance lowers the risk score (\(-1.5\)), while late submissions raise it (\(+2.0\)).

Computing \(Xw\) row by row:

\begin{itemize}[leftmargin=*]
\item \textbf{Student 1:} \((-1.5)(0.90) + (2.0)(1) = -1.35 + 2.0 = 0.65\)
\item \textbf{Student 2:} \((-1.5)(0.60) + (2.0)(4) = -0.90 + 8.0 = 7.10\)
\item \textbf{Student 3:} \((-1.5)(0.80) + (2.0)(0) = -1.20 + 0.0 = -1.20\)
\end{itemize}

\[
Xw =
\begin{bmatrix}
0.65 \\
7.10 \\
-1.20
\end{bmatrix}.
\]

Student 1's weighted score is \(0.65\). On its own that number is just a linear score; a later policy would still need thresholds or calibration to translate it into categories such as low, moderate, or high risk. Student 2's score of \(7.10\) is larger because four late submissions dominate. Student 3's score of \(-1.20\) is smaller because solid attendance with no late submissions pushes the score downward.

The key idea is that the same weight vector \(w\) is applied to every row at once. The matrix does not change the structure of the computation---it only packages the computation so every student is scored in one operation.
\end{example}

\section*{Sparse, Dense, and Missingness as Representation Facts}

Representations are not only numeric. They differ in density, noise, and missingness, and those differences change how models behave.

A \emph{sparse} representation has many zero coordinates. A bag-of-words vector for one email drawn from a vocabulary of $50{,}000$ words usually has fewer than $200$ nonzero entries because most vocabulary words simply do not appear in that message. Sparsity is not a deficiency; it is information. A zero coordinate means the corresponding feature is absent.

A \emph{dense} representation has very few zeros. A learned word embedding of dimension $300$ typically populates all $300$ coordinates with small but nonzero values. Dense representations are compact and support smooth geometric operations like cosine similarity, but they offer no inherent interpretability for any individual coordinate.

A \emph{partially observed} representation has coordinates missing for structural reasons, not random ones. A student record may have quiz scores for seven of ten weeks because the student joined late. A sensor trace may be missing every other reading because the device transmits on an alternating schedule. In each case, the absence of a value carries information about the process that generated the data.

\begin{table}[t]
\centering
\small
\setlength{\tabcolsep}{4pt}
\begin{tabular}{p{1.89cm}p{3.26cm}p{4.8cm}}
\toprule
Type & Example & What absence means \\
\midrule
Sparse & Bag-of-words email vector & Word did not appear in this message \\
Dense & Learned word embedding & Every coordinate is informative; no zero convention \\
Partially observed & Student engagement record with gaps & Student may not have been enrolled yet, or event did not fire \\
\bottomrule
\end{tabular}
\caption{Representation quality includes density, which coordinates are absent, and whether absence is informative. A zero can mean ``not present,'' ``missing by design,'' or ``unknown,'' and those meanings matter for modeling.}
\label{tab:sparse-dense-missing}
\end{table}

The same written value can therefore mean different things. If a quiz-score feature is recorded as \(0\), the representation may be saying one of three incompatible statements:

\begin{center}
\small
\begin{tabular}{p{2.1cm}p{6.9cm}}
\toprule
Encoding & Claim the representation makes \\
\midrule
\(0\) as score & The student took the quiz and earned no points \\
\(\mathrm{NA}\) or missing & The system does not know the score, or the quiz was not applicable yet \\
Imputed \(0\) & The original value was missing, but the model is being shown a zero plus, ideally, a missingness flag \\
\bottomrule
\end{tabular}
\end{center}

Those encodings are not interchangeable. Treating unknown as zero can teach a model that the absence of a record is the same as poor performance.

Three practical lessons follow.

\emph{Sparse representations support fast inner products but can break distance-based methods.} If most coordinates of two vectors are zero, their Euclidean distance is dominated by the dimensions where both are nonzero---or by the few nonzero dimensions where only one is present. Cosine similarity sometimes handles this better, but neither operation treats ``both missing this word'' and ``both present on this word'' the same way.

\emph{Dense embeddings lose interpretability at the coordinate level.} A single coordinate of a word embedding does not mean anything human-readable. The space as a whole may encode meaning geometrically; the individual coordinates do not.

\emph{Missingness in tabular data is often not random.} Imputing a missing value with a mean or zero treats the missing entry as if it were typical, absent, or otherwise uninformative under the chosen coding. But if the value is missing because the student dropped out before week eight, or because a sensor was offline after a power failure, the imputation injects a different kind of distortion. Later chapters on reliability and deployment return to this point. For now, the lesson is that ``missing'' is a value in the representation, not an error to be erased before modeling.

\section*{Geometry: Distance, Similarity, and Scale}

Once cases live in vectors, geometry starts to matter. The geometry is not a picture drawn after the fact; it is part of the model.

If two represented students are
\[
x =
\begin{bmatrix}
0.92 \\
0.80 \\
0.10 \\
0.35
\end{bmatrix},
\qquad
z =
\begin{bmatrix}
0.88 \\
0.78 \\
0.12 \\
0.40
\end{bmatrix},
\]
their Euclidean distance is
\begin{align*}
\|x-z\|_2 &= \bigl[(0.92{-}0.88)^2 + (0.80{-}0.78)^2 \\
&\quad\;+ (0.10{-}0.12)^2 + (0.35{-}0.40)^2\bigr]^{1/2}.
\end{align*}
This distance is small, so the two cases are close under that geometry.

Cosine similarity asks a different question. Instead of asking whether the vectors are close in absolute position, it asks whether they point in nearly the same direction:
\[
\cos(\theta) = \frac{x^\top z}{\|x\|_2 \|z\|_2}.
\]
This is useful when overall scale should matter less than pattern. In language or embedding-style problems, two long vectors can represent similar meaning even when one is globally larger than the other.

\begin{example}[Distance and Cosine Can Rank Neighbors Differently]
Suppose a short document is represented by counts of two words:
\[
q = \begin{bmatrix}1\\1\end{bmatrix},\qquad
a = \begin{bmatrix}2\\2\end{bmatrix},\qquad
b = \begin{bmatrix}1\\3\end{bmatrix}.
\]
Euclidean distance gives
\[
\|q-a\|_2=\sqrt{2}, \qquad \|q-b\|_2=2,
\]
so \(a\) is closer. Cosine similarity gives
\[
\cos(q,a)=1,\qquad \cos(q,b)=\frac{4}{\sqrt{2}\sqrt{10}}\approx 0.89.
\]
Cosine still prefers \(a\), but for a different reason: \(a\) points in exactly the same direction as \(q\), even though it is larger in magnitude. If a third candidate were \(c=[10,10]^\top\), Euclidean distance would call it far away while cosine would call it identical in direction. The right geometry depends on whether amount or pattern is what matters.
\end{example}

\begin{failurebox}
Distance-based methods can fail for a boring reason: one coordinate dominates because it lives on a much larger numeric scale. If age is stored in years but income in dollars, raw Euclidean distance mostly measures income. That is not a property of the world; it is a property of the chosen units.
\end{failurebox}

\section*{Norms, Size, and Regularization Intuition}

A norm measures the size of a vector. Two common norms are
\[
\|w\|_1 = \sum_{j=1}^{d} |w_j|,
\qquad
\|w\|_2 = \sqrt{\sum_{j=1}^{d} w_j^2}.
\]

Later chapters use these ideas for regularization. The reason is simple: a model with very large coefficients tends to be brittle, overly reactive, and too willing to fit accidental quirks in the training data. Penalizing coefficient size expresses a preference for simpler behavior unless the data strongly demands otherwise.

The word \emph{regularization} should mean more than ``extra term in the objective.'' The useful modeling question is:

\begin{quote}
How much coefficient size, complexity, or sensitivity are we willing to tolerate before the fit becomes unstable?
\end{quote}

\section*{Linear Maps, Projections, and Fitting}

A linear map is a rule that transforms one vector into another while preserving weighted-sum structure. In AI terms, it is a structured feature transformation.

This matters because many models do not act directly on the raw input. They act on transformed coordinates. Even when later chapters discuss embeddings or learned internal layers, one recurring question survives:

\begin{quote}
What transformation makes the relevant structure easier to see?
\end{quote}

Projection is the cleanest first fitting intuition. When the world is more complicated than the model class we have allowed, training is often not about finding the truth exactly. It is about finding the closest allowed approximation under a chosen notion of error.

That is the useful least-squares picture. A linear model does not win by capturing every detail of reality; it wins when the best projection into the linear family is already good enough for the task.

\begin{mathlens}
For the simplest projection onto a one-dimensional direction \(u\) with \(\|u\|_2=1\), the fitted point is
\[
\hat{x}=(x^\top u)u.
\]
The residual is \(r=x-\hat{x}\). The key geometric fact is
\[
r^\top u=0,
\]
so the leftover error is orthogonal to the direction we kept. In least squares, this is the meaning of ``best fit'': after projection, no remaining error points along the allowed direction.
\end{mathlens}

\begin{mathlens}
A projection map is itself a linear map. If \(u\) is a unit direction, then
\[
P_u(x)=(u^\top x)u
\]
keeps only the component of \(x\) that lies along \(u\). Applying it twice changes nothing:
\[
P_u(P_u(x))=P_u(x).
\]
That is the algebraic version of geometric fitting: once a vector has been projected onto the allowed line, projecting again cannot improve it further.
\end{mathlens}

\begin{advancednote}
Students often hear that linear models are ``simple'' and stop there. A better statement is that linear models define a restricted hypothesis class with very transparent geometry. The restriction is sometimes a weakness, but it is also what makes the class interpretable, diagnosable, and often surprisingly competitive.
\end{advancednote}

\begin{example}[Projection in One Dimension]
Suppose we observe three data points \(y = [2, 4, 6]\) and we want to fit a constant model---a single number \(c\) that stands in for all three values. Which constant is best?

The mean is \(\bar{y} = (2 + 4 + 6)/3 = 4\). The residuals from the mean are
\[
r = [2 - 4,\; 4 - 4,\; 6 - 4] = [-2,\; 0,\; 2].
\]
The sum of squared residuals is
\[
SS(\bar{y}) = (-2)^2 + 0^2 + 2^2 = 4 + 0 + 4 = 8.
\]

Now try a different constant, say \(c = 3\):
\[
r = [2-3,\; 4-3,\; 6-3] = [-1,\; 1,\; 3].
\]
\[
SS(3) = (-1)^2 + 1^2 + 3^2 = 1 + 1 + 9 = 11.
\]

The sum of squares rose from \(8\) to \(11\) when we moved away from the mean. Any other constant produces a sum of squares no smaller than \(8\). In coefficient terms, the best constant is \(4\). In geometric terms, the data vector \([2,4,6]\) projects onto the constant-model class as \([4,4,4]\), the unique member of that class that sits closest to the data under squared error.

This is the geometry of least-squares fitting in miniature. The model class---all constant functions---is a restricted family. Training finds the member of that family closest to the target values. The residuals \([-2, 0, 2]\) are orthogonal to the constant direction: they average to zero, so no constant shift can reduce the error further.
\end{example}

\section*{Linear Separability and What a Hyperplane Can Actually Express}

A linear score $w^\top x + b$ divides the input space by the boundary $w^\top x + b = 0$. One side has positive score, the other negative score, and points with score exactly zero lie on the boundary itself. In two dimensions, that boundary is a line; in three dimensions, a plane; in $d$ dimensions, a hyperplane.

Understanding what a hyperplane can and cannot separate gives a concrete reason for why later chapters need trees, kernels, and neural nonlinearities. Without this picture, ``nonlinearity matters'' sounds like folklore. With it, the limitation becomes a specific geometric fact.

\textbf{A case where a line succeeds.} Consider four points in two dimensions:
\[
(1,1),\ (1,2),\ (3,3),\ (4,4).
\]
Suppose the first two are class A and the last two are class B. A line such as $x_1 + x_2 = 5$ separates them cleanly. Here the linear model is sufficient.

\textbf{A case where a line fails: the XOR pattern.} Now consider four points:
\[
(0,0),\ (1,1),\ (1,0),\ (0,1).
\]
Suppose $(0,0)$ and $(1,1)$ are class A, and $(1,0)$ and $(0,1)$ are class B. Any line through this plane will put at least one point on the wrong side. There is no linear decision boundary that correctly separates these four points.

\begin{figure}[t]
\centering
\resizebox{\linewidth}{!}{%
\begin{tikzpicture}[font=\small]
\begin{scope}[x=0.75cm, y=0.75cm]
\draw[->] (-0.2,0) -- (4.6,0) node[right] {$x_1$};
\draw[->] (0,-0.2) -- (0,4.6) node[above] {$x_2$};
\filldraw (1,1) circle (2pt) node[below left] {A};
\filldraw (1,2) circle (2pt) node[above left] {A};
\draw (3,3) circle (2pt) node[above right] {B};
\draw (4,4) circle (2pt) node[above right] {B};
\draw[thick, dashed] (0.6,4.4) -- (4.4,0.6);
\node at (2.5,-0.55) {separable by $x_1+x_2=5$};
\end{scope}
\begin{scope}[xshift=5.3cm, x=2.0cm, y=2.0cm]
\draw[->] (-0.2,0) -- (1.8,0) node[right] {$x_1$};
\draw[->] (0,-0.2) -- (0,1.8) node[above] {$x_2$};
\filldraw (0.0,0.0) circle (2pt) node[above right] {A};
\filldraw (1.0,1.0) circle (2pt) node[above right] {A};
\draw (1.0,0.0) circle (2pt) node[above right] {B};
\draw (0.0,1.0) circle (2pt) node[above right] {B};
\node at (0.5,-0.35) {not separable (XOR)};
\end{scope}
\end{tikzpicture}%
}
\caption{Left: a line separates the two classes. Right: the XOR pattern cannot be separated by any line. The filled points are class A; the open points are class B.}
\label{fig:linear-separability}
\end{figure}

The XOR pattern is not a contrived puzzle. It is the general situation where the decision depends on an \emph{interaction} between features: $(0,0)$ and $(1,1)$ belong to the same class not because of either coordinate alone, but because of their combination. A linear model working on raw coordinates cannot represent that.

This is the geometric reason interaction features, kernel methods, tree splits, and neural nonlinearities matter. Each extends the expressible boundary beyond a flat hyperplane: trees by recursive axis-aligned cuts, kernel methods by implicitly working in a richer feature space, and neural networks by composing nonlinear transformations.

\begin{advancednote}
The formal study of which problems linear models can solve falls under the theory of VC dimension and linear dichotomies. For this book, the geometric picture is enough: a hyperplane is a powerful tool when the relevant decision boundary is roughly flat in the feature space, and a poor tool when the boundary requires feature interactions.
\end{advancednote}

\section*{Low-Dimensional Structure}

Real datasets often have many coordinates but fewer genuinely important directions. In image, language, and tabular problems alike, some features overlap, some move together, and some mostly repeat what others already say.

The practical lesson is not that every dataset can be compressed aggressively. It is that high ambient dimension does not always mean high effective complexity. This intuition later supports PCA, embeddings, representation learning, and transfer learning.

\begin{mathlens}
A projection onto an allowed subspace is a linear map. For a unit direction \(u\), define
\[
P_u(x)=(u^\top x)u.
\]
Then \(P_u\) keeps only the component of \(x\) that lies along \(u\). Applying it twice does nothing new:
\[
P_u(P_u(x))=P_u(x).
\]
That idempotence is the algebraic version of geometric compression: once the allowed direction has been kept, projecting again does not change the answer.
\end{mathlens}

A full singular-value-decomposition treatment is not needed here. What students do need is the idea that useful structure can live in fewer directions than the raw coordinate count suggests.

\subsection*{Eigendecomposition and SVD: A Preview}

\begin{advancednote}
Later chapters (dimensionality reduction in Chapter~\ref{ch:structure-beyond-linearity}, embeddings in Chapter~\ref{ch:representation-foundation-models}, and latent-factor models in Chapter~\ref{ch:probabilistic-models}) use the \emph{singular value decomposition} (SVD) and eigendecomposition. A full treatment is not needed yet. The useful bridge idea is only this: data often varies more along some directions than others, and a low-dimensional summary tries to keep the directions that carry the most structure.

\textbf{Eigenvectors and eigenvalues.} For a square matrix $A$, an eigenvector $v$ satisfies $Av = \lambda v$: the matrix stretches $v$ by a factor $\lambda$ (the eigenvalue) without changing its direction. For a covariance matrix, eigenvectors are principal axes; ordered by eigenvalue, larger eigenvalues correspond to more variance along that direction.

\textbf{Singular Value Decomposition.} Any matrix $X$ (not just square ones) can be decomposed as $X = U \Sigma V^\top$, where $U$ and $V$ are orthogonal matrices and $\Sigma$ is diagonal with non-negative entries (the singular values). The columns of $V$ give orthogonal directions ordered by how much squared norm of the matrix they capture. When $X$ is centered first, those directions become the principal-component directions of greatest variance. PCA is therefore computed from the SVD of a centered data matrix: the top $k$ columns of $V$ are the first $k$ principal components.

\textbf{Why this matters for ML.} Truncating the SVD to the top $k$ components gives the best rank-$k$ approximation of the matrix (in the Frobenius norm sense). This is why PCA works for dimensionality reduction, why matrix factorization works for recommendation (Chapter~\ref{ch:ranking-recommendation}), and why low-rank approximations appear throughout representation learning. The key intuition: most of the ``important'' structure in a matrix can often be captured by a small number of directions.

Given a centered data matrix $X \in \mathbb{R}^{n \times d}$ (rows are examples, columns are features), the SVD yields $X = U \Sigma V^\top$. The top \(k\) columns of \(V\) give the first \(k\) principal-component directions. If the first five directions explain most of the centered variance, the data can often be summarized in five dimensions even when it has many raw features. Treat this as a preview, not as a prerequisite for the rest of the bridge.
\end{advancednote}

\section*{Chapter Artifact: Representation Card}

\begin{quote}
\noindent\textbf{Representation Card}

\medskip
\begin{tabular}{>{\raggedright\arraybackslash}p{2.6cm}>{\raggedright\arraybackslash}p{5.4cm}}
\toprule
Field & What to write \\
\midrule
One example is... & Name the basic case: one student, one email, one image patch, one customer session, one hour of sensor data \\
Coordinates are... & Name each feature family and what each coordinate means \\
What is lost? & State what the representation throws away or blurs \\
What geometry matters? & Distance, angle, overlap, ordering, or something else \\
What does a linear score mean? & Explain what a weighted sum would count as evidence for or against \\
Main representation risk & State the most likely mismatch between representation and task \\
\bottomrule
\end{tabular}
\end{quote}

If this card cannot be written clearly, later model debates are usually premature.

\begin{example}[Diverse Representation: Spam Email Filtering]
The task is to decide whether an incoming email is spam. Each row of the dataset is one email. The coordinates are chosen features: word frequencies from a bag-of-words model, a sender reputation score, the number of links in the message, and header flags (all-caps subject, suspicious domains, and so on).

With a vocabulary of 50 words and 5 numeric features (reputation, link count, and three binary flags), the representation lives in $\mathbb{R}^{55}$. The first 50 coordinates are term frequencies; coordinates 51 through 55 are the numeric metadata.

A critical note on representation choice: the bag-of-words representation discards word order entirely. ``Nigerian prince has offer'' and ``offer has prince Nigerian'' both map to the same vector because only counts matter, not position. This is already a modeling decision: we are saying that detecting spam does not require understanding syntax or sequence, only statistics of presence and absence. That choice is often reasonable for spam (many phishing emails repeat the same words in any order), but it fails for sentiment analysis (``not good'' means something very different from ``good not''). Filling out a representation card before choosing a model family makes such assumptions visible.
\end{example}

\begin{example}[Representation Card Row for an Image Patch]
For a simple image task, one example might be a \(32 \times 32\) image patch. The raw coordinates are pixel intensities, with three color channels per pixel. What is lost depends on preprocessing: resizing may blur small objects, cropping may remove context, and converting to grayscale discards color. The geometry matters locally---nearby pixels and small translations are meaningful, while arbitrary coordinate swaps are not. A linear score on raw pixels mostly detects fixed templates; a convolutional model, introduced later, can instead ask repeated local questions across the image.
\end{example}

\section*{Common Mistakes}

\begin{mistakebox}
\item \textbf{Treating a vector as just a bag of numbers.} A vector is a chosen representation, not a random pile of values.
\item \textbf{Confusing row count with feature count.} One case and one feature family play different roles in a dataset.
\item \textbf{Ignoring scale when distance or nearest neighbors matter.} Geometry depends on units.
\item \textbf{Treating a large dot product as self-explanatory.} A weighted sum only matters relative to the chosen representation.
\item \textbf{Thinking that more dimensions automatically mean more useful information.} Extra coordinates can just as easily add noise or redundancy.
\end{mistakebox}

\section*{Looking Ahead}

This bridge supplied the geometric language of representation: vectors, matrices, weighted scores, distance, and projection. The next bridge adds the uncertainty language that makes those scores trainable and interpretable---probability, loss, empirical risk, and gradient signals. Together they prepare the main text, where Chapter 1 turns these mathematical objects into a defensible learning problem.

The SVD preview above gives just enough eigendecomposition intuition for PCA, embeddings, and matrix factorization in later chapters. The full machinery returns when those topics demand it.

\begin{chaptersummary}
\item Linear algebra matters in AI because representation is already part of the model.
\item A vector is one represented case, and a matrix is many represented cases organized together.
\item A dot product is a weighted evidence summary.
\item Distance, similarity, and scale are modeling assumptions, not just visualization choices.
\item Norms measure size and later become regularization tools.
\item The main deliverable of the chapter is a representation card written before declaring that a model family is a good fit.
\end{chaptersummary}

\begin{table}[h]
\centering
\small
\caption{Where linear algebra ideas return in later chapters.}
\begin{tabular}{@{}L{2.75cm}L{7.05cm}@{}}
\toprule
\textbf{Bridge Topic} & \textbf{Used In} \\
\midrule
Dot products and weighted sums & Chapters~\ref{ch:linear-models},~\ref{ch:optimization-neural-networks} (linear models, neural networks) \\
Norms and distance & Chapters~\ref{ch:linear-models},~\ref{ch:structure-beyond-linearity},~\ref{ch:representation-foundation-models} (regularization, nearest neighbors, embeddings) \\
Cosine similarity & Chapters~\ref{ch:representation-foundation-models},~\ref{ch:language-grounding} (embeddings, retrieval) \\
Matrix--vector products & Chapter~\ref{ch:optimization-neural-networks} (forward pass, backpropagation) \\
Linear separability & Chapters~\ref{ch:linear-models},~\ref{ch:optimization-neural-networks} (baselines, learned features) \\
\bottomrule
\end{tabular}
\end{table}

\begin{studyquestions}
\item Why is representation choice already a modeling decision rather than a neutral preprocessing step?
\item In plain language, what does the dot product \(w^\top x\) mean in a decision-support problem?
\item Why can Euclidean distance and cosine similarity lead to different judgments about what counts as ``similar''?
\item What does it mean to say that fitting can be understood as projection into an allowed model class?
\end{studyquestions}

\begin{exercises}
\item \exercisetag{Design}{Core}Write a four-coordinate vector for a student-support problem and explain the meaning of each coordinate in one sentence.
\item \exercisetag{Concept}{Core}For the vector \(x = [0.9, 0.7, 0.2]^\top\) and the weight vector \(w = [-2, 1, 3]^\top\), compute \(w^\top x\) and explain what each contribution means.
\item \exercisetag{Design}{Core}Construct a \(3 \times 4\) design matrix for three emails with four hand-chosen features. State clearly what the rows and columns represent.
\item \exercisetag{Concept}{Core}Give one situation in which cosine similarity would be a better comparison than Euclidean distance, and one situation in which it would be worse.
\item \exercisetag{Concept}{Intermediate}Fill out a representation card for spam filtering, student support, or a simple image task. If you plan to carry one running case through the book, keep the same case and add one sentence saying which coordinates Chapter 1's framing memo would treat as acceptable inputs at prediction time.
\end{exercises}

\begin{challengeexercises}
\item \exercisetag{Concept}{Advanced}Show with a tiny numeric example that rescaling one feature can change which case is the nearest neighbor, even when all other coordinates stay the same.
\item \exercisetag{Concept}{Advanced}Give two different representations for the same task, explain what each preserves, and argue which one makes a linear score more meaningful.
\item \exercisetag{Concept}{Advanced}Suppose a target cannot be expressed exactly by any weighted sum of the available features. Explain why least-squares fitting can still be read as finding the closest allowed predictor.
\end{challengeexercises}
