\chapter*{Mathematical Bridge II: Probability, Risk, and Learning Signals}
\addcontentsline{toc}{chapter}{Mathematical Bridge II: Probability, Risk, and Learning Signals}

A model outputs the number $0.83$. Is that a probability? A score? A confidence? Combining it with another such number---do you multiply, sum, or take the log? Each training chapter that follows assumes a settled vocabulary for reading numbers like that one: score, probability, decision rule, pointwise loss, dataset-level objective. This bridge supplies it.

By the end, the reader should be able to distinguish a score, a probability, a decision rule, a pointwise loss, and a dataset-level training objective; read conditional probability and expectation in plain language before touching formulas; explain why logarithms, exponentials, sigmoid-style transformations, and cross-entropy appear naturally in ML; treat empirical risk as the average training signal over a dataset rather than an abstract formula to memorize; and read gradients and the chain rule as credit-assignment tools for reducing training loss. The functional view of predictors and risks, along with the generative/discriminative contrast and priors and MAP as preview vocabulary, is second-pass material. It is useful for later chapters but not required before continuing.

\begin{prereqbridge}
Use this chapter as just-in-time review. Read it now if probability or training-signal notation feels thin, or return when later chapters start leaning on loss, calibration, and gradients. The chapter assumes comfort with fractions, ratios, and the idea of an average. Familiarity with logarithms helps; we introduce them with worked examples. Calculus (derivatives) appears only in a few short derivations, especially the gradient section, and we explain it from scratch when it appears. Students who have already taken probability should still read the chapter. The AI-specific vocabulary---loss, empirical risk, gradient as credit signal---differs from the standard probability course framing.
\end{prereqbridge}

\begin{practicalnote}
\textbf{Core path.} On a first pass, focus on scores, probabilities, decisions, losses, empirical risk, and gradients as training signals. The functional viewpoint, the generative/discriminative contrast, priors, and MAP are orientation material---skim them until later chapters need them.
\end{practicalnote}

Most students meet probability and optimization in fragments. One course teaches conditional probability. Another introduces derivatives. Then a machine-learning lecture writes down log loss, empirical risk, and gradient descent as if the connection were obvious. This chapter slows that chain down.

The goal is not to flood the reader with formalism. It is to make later training objectives readable. When a later chapter writes down a loss, a score, a calibration curve, or a gradient update, the student should be able to answer three questions immediately:

\begin{itemize}[leftmargin=*]
\item What uncertain quantity is being modeled?
\item What mistake is being penalized?
\item What signal tells the parameters how to move?
\end{itemize}

Separate five objects that students collapse together: the model score, the probability interpretation of that score, the decision rule that acts on it, the pointwise loss on one example, and the dataset-level objective minimized during training.

\textbf{Loss as learning signal.} A loss is not the final judgment of the system. It is the message training uses to say ``this prediction behaved well'' or ``this prediction behaved badly.'' Interpret the message before optimizing it.

Keep the chapter artifact in view from the start. As later sections separate score, probability, decision rule, pointwise loss, and dataset objective, the goal is to fill a \emph{learning-signal card}---not to memorize disconnected formulas.

\section*{Opening Question: What Does A Score Like 0.8 Mean?}

Suppose a model outputs \(0.8\) for a student, an email, or a medical case. That number could mean several different things.

\begin{itemize}[leftmargin=*]
\item An uncalibrated score: larger means riskier, but the number is not yet an honest probability.
\item A probability estimate: about \(80\%\) of comparable cases are expected to be positive.
\item Input to a decision rule: act when the number is above some threshold.
\end{itemize}

Three different interpretations. Many modeling mistakes begin when they get treated as the same thing.

This bridge chapter therefore starts with language. A model output is not automatically a trustworthy probability, and a probability is not automatically a decision.

\section*{Conditional Probability In Plain Language}

Machine-learning prediction asks about uncertainty \emph{given} observed information. The notation
\[
P(y=1 \mid x)
\]
reads as ``the probability that the target is positive given the observed input \(x\).''

The vertical bar means ``given.'' That is the central operational idea.

\begin{itemize}[leftmargin=*]
\item In spam filtering, \(x\) might be the message text and metadata.
\item In student support, \(x\) might be attendance, quizzes, and submissions.
\item In image classification, \(x\) might be the visual measurement.
\end{itemize}

Conditional probability matters because AI systems do not predict labels in a vacuum. They predict them given whatever evidence the representation makes available.

If the notation feels dense, translate it into a sentence: ``given the evidence we observed, how plausible is the target event?'' That sentence is the main idea. The notation is just the compact version.

\section*{Expectation As Average Consequence}

Expectation is the language of average consequence. When a quantity varies, its expectation summarizes the average value across repeated comparable cases.

This idea appears constantly in AI:

\begin{itemize}[leftmargin=*]
\item average loss over a training set,
\item average reward in decision-making,
\item average error over future unseen cases,
\item average outcome under a probabilistic forecast.
\end{itemize}

Suppose a support team can contact one student. Contacting a truly struggling student creates \(+5\) units of benefit; contacting a student who did not need help costs \(-1\) unit of attention. If a model estimates a \(0.8\) chance that the student truly needs help, the expected value of contacting is
\[
0.8(5) + 0.2(-1) = 4 - 0.2 = 3.8.
\]

This calculation does not claim that life has units called \(3.8\). It shows how uncertain outcomes become decision-relevant through expected consequence.

\subsection*{Expectation Under A Distribution}

The student-contact calculation above quietly used a recipe that returns throughout the book. Whenever a quantity \(g(Y)\) depends on a random outcome \(Y\), and the outcome takes value \(y\) with probability \(p(y)\), the expectation is the probability-weighted sum (or integral) of \(g\) over outcomes:
\[
\mathbb{E}_{Y \sim p}[g(Y)] \;=\; \sum_y g(y)\,p(y).
\]
The subscript \(Y \sim p\) reads ``\(Y\) drawn from the distribution \(p\).'' It is not a new operation; it is a reminder of \emph{which} distribution the average is taken under.

\textbf{Expectation under a distribution.} Three readings of the same formula recur in later chapters.
\begin{itemize}[leftmargin=*]
\item \emph{Loss over training data.} $\mathbb{E}_{(x,y) \sim \mathcal{D}}[\ell(y, f_\theta(x))]$: average pointwise loss when $(x,y)$ is drawn from the data distribution $\mathcal{D}$. The empirical version replaces the expectation by the sample average.
\item \emph{Reward under a policy.} $\mathbb{E}_{\tau \sim \pi}[R(\tau)]$: average return when trajectories $\tau$ are generated by following a policy $\pi$ (Chapter~\ref{ch:reinforcementlearning}).
\item \emph{Marginal likelihood.} $\mathbb{E}_{z \sim q}[\log p(x, z)]$: average log joint when a latent $z$ is drawn from a variational distribution $q$ (Chapter~\ref{ch:probabilistic-models}, Chapter~\ref{ch:generative-representations}).
\end{itemize}
A change in subscript usually means a change in which world the average is taken under. Two expectations with the same integrand but different subscripts are not the same number, and disagreements between them are exactly what importance sampling, off-policy evaluation, and the ELBO derivation are about.

\subsection*{Jensen's Inequality, In One Line}

Some functions are bowl-shaped (convex), others are dome-shaped (concave). The most useful single fact about expectations and concave functions appears repeatedly when we manipulate log-likelihoods.

\textbf{Jensen's inequality (concave version).} For any concave function \(\varphi\) and any random variable \(X\),
\[
\varphi\bigl(\mathbb{E}[X]\bigr) \;\ge\; \mathbb{E}\bigl[\varphi(X)\bigr].
\]
For a convex function the inequality reverses. The function we use most is \(\varphi = \log\), which is concave, so
\[
\log\bigl(\mathbb{E}[X]\bigr) \;\ge\; \mathbb{E}[\log X]
\]
whenever \(X > 0\). Geometric reading: the chord lies below the curve for concave functions, so averaging inputs and then applying the function beats applying the function and then averaging.

\textbf{Where Jensen returns.} Chapter~\ref{ch:probabilistic-models} uses Jensen to show that the EM algorithm cannot make the log-likelihood worse. Chapter~\ref{ch:generative-representations} uses the same inequality to derive the evidence lower bound (ELBO) that VAE training maximizes: the bound is exactly the slack in Jensen's inequality applied to \(\log p(x) = \log \mathbb{E}_{z \sim q}[p(x \mid z) p(z) / q(z)]\). When a later chapter writes ``by Jensen,'' read the line above and it will not feel like sleight of hand.

\subsection*{Multivariate Gaussians, Without The Heavy Notation}

The univariate Gaussian density \(\mathcal{N}(x \mid \mu, \sigma^2)\) is a familiar bell curve over the real line: it puts most of its mass near the mean \(\mu\) and decays at a rate set by the standard deviation \(\sigma\). The multivariate version replaces the mean with a vector and the variance with a matrix, and that is essentially all that is new.

\textbf{The multivariate Gaussian.} For \(x \in \mathbb{R}^d\), mean \(\mu \in \mathbb{R}^d\), and a symmetric positive-definite covariance matrix \(\Sigma \in \mathbb{R}^{d \times d}\), the density is
\[
\mathcal{N}(x \mid \mu, \Sigma) \;=\; \frac{1}{(2\pi)^{d/2}\,|\Sigma|^{1/2}}\,\exp\!\Bigl(-\tfrac{1}{2}(x-\mu)^\top \Sigma^{-1}(x-\mu)\Bigr).
\]
Three readings make the formula tractable:
\begin{itemize}[leftmargin=*]
\item \emph{Mean} \(\mu\) is the centre of the cloud.
\item \emph{Covariance} \(\Sigma\) is a matrix of variances and covariances: diagonal entries give the spread along each coordinate; off-diagonal entries say which coordinates move together.
\item \emph{Mahalanobis distance} \((x-\mu)^\top \Sigma^{-1}(x-\mu)\) generalizes ``how many standard deviations away'' to a coordinate system that respects covariance. The exponent is just \(-\tfrac{1}{2}\) times that distance.
\end{itemize}
When \(\Sigma\) is diagonal, the density factorizes into a product of univariate Gaussians on each coordinate. The level sets of the density are ellipsoids; the shape and orientation are exactly the eigenstructure of \(\Sigma\).

The multivariate Gaussian is the only continuous distribution most of this book actually uses to model latent or noise structure. It returns in Chapter~\ref{ch:probabilistic-models} (Gaussian mixture models, Bayesian linear regression, Gaussian processes) and Chapter~\ref{ch:generative-representations} (the prior over latent codes in a variational autoencoder, and the noise schedule in diffusion models). When a later chapter writes \(\mathcal{N}(x \mid \mu, \Sigma)\), the formula above is what it means; the rest is mostly bookkeeping.

\section*{Base Rates, Bayes Decisions, and Cost-Sensitive Action}

Probability and action are different things. A model can produce a well-calibrated probability estimate, yet the right action can still differ across teams, deployments, and time periods, even when the probability stays the same.

The bridge from probability to action runs through three factors: the estimated probability, the costs of the available actions, and the prevalence and workload context of deployment.

Consider the student-support example. Suppose the model assigns probability $0.35$ to a student disengaging. Is that high enough to trigger an outreach call?

The probability alone cannot answer the question. The answer depends on:
\begin{itemize}[leftmargin=*]
\item the benefit of a correct intervention---large if it prevents serious academic harm,
\item the cost of a false alarm---staff time and student inconvenience,
\item the prevalence and workload context. If $30\%$ of students routinely score near $0.35$, routine outreach may be sustainable. If only $2\%$ ever reach that level, the same policy creates a very different workload.
\end{itemize}

\begin{example}[Same Probability, Different Action]
A medical triage team and a marketing team both receive a model score of $0.35$. The medical team escalates because missing a true case has severe consequences and resources are available. The marketing team does not act because their cost-per-contact is high relative to the expected value of conversion at $35\%$ likelihood. Same probability, different correct action. The model output did not contain the action. The action required a cost structure on top of the probability.
\end{example}

Bayes-optimal decisions formalize this idea. In the simplest binary setup, assume the model score is already the calibrated posterior probability of the positive class, correct decisions carry no extra utility beyond avoiding mistakes, and the relevant costs are $C_{FP}$ (false positive) and $C_{FN}$ (false negative). The optimal threshold $\tau$ then satisfies
\[
\tau = \frac{C_{FP}}{C_{FP} + C_{FN}}.
\]
The derivation is just an expected-cost comparison. If the posterior probability of the positive class is \(p\), predicting positive has expected mistake cost \((1-p)C_{FP}\), and predicting negative has expected mistake cost \(pC_{FN}\). Predict positive when
\[
(1-p)C_{FP} < pC_{FN}.
\]
Solving this inequality gives the threshold above. If false negatives cost nine times more than false positives, the optimal threshold is \(1/(1+9)=0.1\), not \(0.5\). The default threshold of \(0.5\) therefore encodes an implicit equal-cost assumption that is often wrong in practice. Base-rate changes still matter, but they enter through the calibrated posterior probabilities the threshold is applied to.

\begin{example}[A Small Bayes Calculation]
Suppose only \(10\%\) of messages in a course-support queue are truly urgent. A keyword such as ``deadline'' appears in \(60\%\) of urgent messages and \(20\%\) of non-urgent messages. If a new message contains that keyword, Bayes' rule gives
\[
P(\text{urgent}\mid\text{keyword}) =
\frac{0.60 \cdot 0.10}{0.60 \cdot 0.10 + 0.20 \cdot 0.90}
= \frac{0.06}{0.24}=0.25.
\]
The keyword raised the probability from \(0.10\) to \(0.25\), but the action still depends on cost and capacity. If urgent misses are very costly, \(0.25\) may justify review. If review slots are scarce and the keyword is common, the same probability may be too low to trigger automatic escalation.
\end{example}

The card should already distinguish the model output from the decision rule. A probability such as \(0.25\) belongs in the model-output row; the outreach threshold belongs in the decision-rule row.

\textbf{Separate three questions.} Does the model output an honest probability? What are the costs of each kind of mistake? What is the current prevalence and score distribution in deployment? Only after all three are answered does it make sense to ask what the action should be.

A common undergraduate mistake is to assume that the best classifier, the best probability forecaster, and the best decision policy are the same object. They are not. A model with strong ranking quality can still produce poorly calibrated probabilities. A well-calibrated model can still imply different thresholds in high-stakes versus low-stakes settings. And both may need further adjustment when prevalence or score distribution shifts---seasonally, demographically, or across deployment contexts.

\section*{Logarithms, Exponentials, and Why They Keep Appearing}

Probabilities multiply. When a sequence of events or labels is modeled jointly, the full likelihood becomes a product of many factors. Logarithms turn products into sums:
\[
\log(ab) = \log a + \log b.
\]

That simple identity is one reason log loss and log-likelihood show up throughout ML. Sums are easier to analyze and optimize than long products.

Exponentials matter for the reverse reason. They turn linear or affine scores into positive quantities that can be normalized into probabilities. The sigmoid function
\[
\sigma(z) = \frac{1}{1+e^{-z}}
\]
maps any real score \(z\) into a number between \(0\) and \(1\). Large positive scores map near \(1\), large negative scores map near \(0\), and \(z=0\) maps to \(0.5\).

If \(p\) is the probability of class \(1\), then the odds are \(p/(1-p)\). Taking logs gives the log-odds \(z=\log\!\bigl(p/(1-p)\bigr)\). Solving for \(p\) gives \(p=e^z/(1+e^z)=1/(1+e^{-z})\), which is exactly the sigmoid.

Why does the sigmoid appear so often? We need a function that maps any real-valued score to $[0, 1]$, producing something that behaves like a probability. The sigmoid is a smooth S-curve that does exactly this: very negative scores map close to 0, very positive scores map close to 1, and the transition is gradual. The logistic function $\sigma(z) = 1/(1 + e^{-z})$ is the simplest such curve, and it arises naturally as the inverse of the log-odds transformation.

Students should not memorize this only as a formula. Read it as a translation rule:

\begin{quote}
a linear score can be turned into a number between \(0\) and \(1\) through a smooth nonlinear map
\end{quote}

In simple binary classifiers, that number often reads as an estimated class probability. That reading is useful, but not magic. It depends on how the model was trained and whether its outputs are later checked for calibration.

\section*{Loss Functions As Learning Signals}

A loss function is a penalty on one prediction. It measures how bad the current output was on one example. Three common losses are worth reading carefully.

Throughout this book, \(\ell(y, \hat{y})\) denotes a loss comparing a true label \(y\) to a generic prediction \(\hat{y}\). When the prediction is specifically a probability, we write \(\ell(y, p)\) to emphasize that \(p \in [0,1]\). The distinction matters: some losses (such as log loss) require a valid probability, not just a raw score.

\begin{itemize}[leftmargin=*]
\item \textbf{Squared error}: \((y-\hat{y})^2\). Large misses are punished strongly.
\item \textbf{Absolute error}: \(|y-\hat{y}|\). Misses grow linearly.
\item \textbf{Log loss / cross-entropy}: penalizes assigning low probability to the true label, especially when the model is confidently wrong.
\end{itemize}

For binary classification with label \(y \in \{0,1\}\) and predicted probability \(p\), the pointwise log loss is
\[
\ell(y,p) = -\bigl[y\log p + (1-y)\log(1-p)\bigr].
\]
If the true label is \(1\), this becomes \(-\log p\). When the model says \(p=0.99\), the penalty is tiny. When it says \(p=0.01\), the penalty is severe. That is why log loss is the standard teaching example for ``confident mistakes are expensive.''

\begin{example}[Computing Cross-Entropy Loss by Hand]
Two models predict the probability that a student is at risk. The true label is \(y=1\) (the student genuinely is at risk).

\begin{itemize}[leftmargin=*]
\item \textbf{Model A} predicts \(p_A = 0.8\).
\item \textbf{Model B} predicts \(p_B = 0.3\).
\end{itemize}

Since \(y=1\), the log loss formula reduces to \(\ell = -\log p\). Computing each:
\[
L_A = -\log(0.8) \approx 0.22,
\qquad
L_B = -\log(0.3) \approx 1.20.
\]

Model A is penalized far less because it was more confident in the correct direction. Model B assigned low probability to what actually happened, so its penalty is steep.

The log makes the penalty grow sharply when the model is confidently wrong. A model that predicts \(p = 0.01\) for a true positive case suffers a loss of \(-\log(0.01) \approx 4.61\)---more than twenty times Model A's penalty. This asymmetry is what makes log loss the standard choice when calibrated probability estimates matter.
\end{example}

The right question is never ``which loss is standard?'' The better question is:

\begin{quote}
What kind of wrong behavior does this loss care about, and what kind does it still permit?
\end{quote}

Squared error reacts strongly to outliers. Absolute error is less sensitive. Log loss cares about confidence. A model that says \(0.99\) and is wrong is punished much more harshly than a model that says \(0.60\) and is wrong.

\begin{center}
\small
\begin{tabular}{p{2.3cm}p{2.77cm}p{4.46cm}}
\toprule
Loss & What it sees & Main modeling message \\
\midrule
Squared error & Numeric discrepancy & Large misses matter much more than small ones \\
Absolute error & Numeric discrepancy & A few very large misses do not dominate as strongly \\
Log loss & Probability assigned to the truth & Confident mistakes are especially serious \\
\bottomrule
\end{tabular}
\end{center}

The formulas above apply to binary classification with $y \in \{0,1\}$. Many tasks have more than two classes. Chapter~\ref{ch:prediction-loss-decision} extends the evaluation story to multiclass confusion matrices and macro versus micro averaging. Chapter~\ref{ch:optimization-neural-networks} introduces the standard logit-and-softmax output layer for multiclass prediction. Internalize the binary version first---it makes every idea (confident mistakes, proper scoring, calibration) visible with the fewest symbols.

\section*{Proper Scoring Rules and Honest Probability Forecasts}

Not every loss encourages a model to report its true beliefs. A \emph{proper scoring rule} makes honest forecasting the safest strategy: if a model genuinely believes the probability of an event is $p$, the expected loss is minimized by reporting exactly $p$ rather than some distorted value.

Log loss is the classic example. Suppose the true probability of a positive outcome is $p^* = 0.7$. A model that reports $\hat{p}$ incurs expected log loss
\[
\mathbb{E}[\ell] = -p^* \log \hat{p} - (1-p^*) \log(1-\hat{p}).
\]
Taking the derivative with respect to $\hat{p}$ and setting it to zero gives $\hat{p} = p^*$. Reporting the true probability minimizes expected log loss. Shading the report in either direction makes things worse.

Accuracy, by contrast, is not a proper scoring rule for probabilities. Under accuracy, any probability above $0.5$ produces the same binary prediction of class 1. A model is not penalized for being overconfident at $0.95$ when the true probability is $0.7$; both produce the same prediction. Accuracy gives no incentive to calibrate.

A scoring rule $S(\hat{p}, y)$ is \emph{proper} if $\mathbb{E}_{y \sim p^*}[S(p^*, y)] \leq \mathbb{E}_{y \sim p^*}[S(\hat{p}, y)]$ for all $\hat{p}$, with equality only when $\hat{p} = p^*$ in the strictly proper case. Log loss and Brier score are proper. Accuracy-based rewards are not. Proper rules are the right losses to use when calibrated probabilities matter for downstream decisions.

Why does this matter for AI practice? When a model output drives a threshold decision, the calibration of that output determines whether the threshold can be set reliably from a stated probability. Consider a model that outputs $0.82$ for cases whose empirical event rate is much lower. Under a threshold-as-policy interpretation, it will flag many cases that a properly calibrated model would not. The team can tighten the threshold empirically, but without recognizing that the underlying probabilities are systematically distorted, they cannot trust the threshold to transfer to a new semester, a new patient cohort, or a new deployment context.

When selecting a training loss, ask whether it rewards honest probability forecasts or merely correct binary predictions. For any application where the probability score will drive a downstream decision, prefer log loss or Brier score over accuracy as the training signal.

\section*{From One Example To Many: Empirical Risk}

What we would ideally minimize is future expected loss on the population of cases the system will actually face. We almost never know that quantity directly. What training can compute is the average loss on the sample in hand---a practical stand-in for the unknown population risk.

If \(f\) is a predictor and \(\mathcal{D}\) is the data-generating process, then the population risk is
\[
R(f) = \mathbb{E}_{(X,Y)\sim \mathcal{D}}\!\bigl[\ell(Y,f(X))\bigr],
\]
while the empirical risk on a sample \((x_i,y_i)_{i=1}^n\) is
\[
\hat{R}_n(f) = \frac{1}{n}\sum_{i=1}^n \ell\bigl(y_i,f(x_i)\bigr).
\]
The sample average is the stand-in because we can compute it now, and if the sample is representative it estimates the future loss we care about.

Training does not optimize one example at a time forever. It optimizes an objective over a dataset. If the model is \(f_\theta\), the pointwise loss is \(\ell\), and the dataset is \((x_i, y_i)_{i=1}^n\), a standard training objective is
\[
\mathcal{L}_{\text{train}}(\theta)
=
\frac{1}{n}
\sum_{i=1}^{n}
\ell\bigl(y_i, f_\theta(x_i)\bigr).
\]

This is empirical risk: the average observed loss on the available sample.

The operational reading is more important than the notation:

\begin{itemize}[leftmargin=*]
\item make a prediction on each training example,
\item score that prediction with a pointwise loss,
\item average the penalties,
\item change the parameters to reduce the average.
\end{itemize}

This is the bridge from one mistake to a trainable objective.

\section*{Population Risk Versus Empirical Risk}

Training computes one quantity. The quantity we actually care about is different.

The \emph{empirical risk} is the average loss over the examples in hand. We can compute it. It changes as we update the parameters, and we minimize it during training.

The \emph{population risk} is the expected loss over all the cases the system will face in deployment: future students, future emails, future images from a new camera. We cannot compute it directly. We have only a sample, and even that sample may not perfectly represent the future.

This gap is one of the deepest ideas in the book. Its consequences reach every chapter:
\begin{itemize}[leftmargin=*]
\item A model that achieves very low training loss can still have high population risk if it has memorized accidents of the training sample.
\item A model that generalizes well has low population risk even when its training loss is not minimal.
\item Validation and test sets are our only window onto population risk---which is why contaminating them with training decisions is so damaging.
\end{itemize}

\begin{example}[Why Low Training Loss Is Not Yet Evidence]
Suppose a model memorizes every training example perfectly: for each student in the training set it has learned an exact rule that reproduces the label. Its training error is zero, and under some losses its training loss may be near zero. But when the model meets a new student---one it has never seen---it has no generalizable rule to fall back on. Its population risk may be high. The perfect training result was a mirage. It said something about the sample but nothing about the distribution.
\end{example}

Treat training loss as a signal about optimization and population risk as the actual goal. When someone says ``the model improved,'' they should mean population risk improved, as estimated by held-out performance. Saying ``training loss went down'' without that translation is not yet evidence of learning in the useful sense.

\textbf{Population risk discipline.} When training loss is offered as evidence of learning, ask: what do the held-out numbers say? If only training loss has been reported, the argument is incomplete.

\section*{A Gentle Functional View Of Loss Functions}

The shift this section asks for: stop looking only at the coordinates of \(\theta\), and start looking at the whole predictor that training is selecting.

A model is not only a parameter vector. It is also a function. Many students never get this shift stated clearly enough.

\textbf{The load-bearing fact.} When the loss $\ell(\hat y, y)$ is convex in $\hat y$, the risk functional $R[f] = \mathbb{E}_{X,Y}[\ell(f(X), Y)]$ is convex in the predictor $f$ in the following operational sense: for any two predictors $f_0, f_1$ and any $\lambda \in [0, 1]$, the mixed predictor $f_\lambda(x) = (1-\lambda) f_0(x) + \lambda f_1(x)$ satisfies
\[
R[f_\lambda] \;\le\; (1-\lambda)\, R[f_0] + \lambda\, R[f_1].
\]
This is the formal reason model averaging and ensembling never hurt expected risk: averaging predictors on a convex loss can only reduce risk, never increase it. The same fact powers Bayesian model averaging and any time you find yourself blending two reasonable models on a calibration-aware loss. The functional view earns its name here---convexity is a property of the rule that scores whole predictors, not of any single parameter coordinate.

\subsection*{A Predictor Is A Function}

When we write \(f_\theta(x)\), the parameter vector \(\theta\) matters, but the real object used at prediction time is the mapping \(x \mapsto f_\theta(x)\). The deployed system receives an input and returns an output. The trained model is a predictor function.

This viewpoint matters because later questions---generalization, regularization, hypothesis-class choice---are easier to think about at the function level than at the parameter level.

\subsection*{Pointwise Loss Versus Risk Functional}

The loss \(\ell(y, \hat{y})\) is pointwise. It looks at one target and one prediction. Empirical risk, by contrast, takes the whole predictor and returns one number summarizing how it behaved across the dataset.

That is the useful light functional-analysis idea in this book:

\begin{quote}
a risk objective is a rule that assigns one scalar value to a whole predictor
\end{quote}

Seen this way, the formulas above are not mere notation changes. One is the ideal quantity defined by the real world; the other is the computable approximation used during training.

Only after the student understands that sentence does the term \emph{functional} become useful.

\subsection*{Hypothesis Classes As Sets Of Allowable Predictors}

Training is not choosing from all imaginable functions. It is choosing from a restricted class:

\begin{itemize}[leftmargin=*]
\item all linear predictors on chosen features,
\item all trees up to some depth,
\item all neural networks of a chosen architecture,
\item all models produced by a certain pretraining and adaptation pipeline.
\end{itemize}

This is the hypothesis class. It matters because every training objective answers the question: which predictor in this allowed family looks best under the chosen risk?

\subsection*{Norms, Complexity, and Regularization}

Regularization becomes easier to interpret once predictors are seen as members of a function class. Penalizing coefficient size is one way of saying:

\begin{quote}
among predictors that fit the data similarly well, prefer the less extreme one
\end{quote}

In linear models this appears as a penalty on coefficient norms. In broader function-class language, it expresses a preference for predictors that are simpler, smoother, or less sensitive according to a chosen measure of size.

When a norm penalty is added to the training objective, the combined loss becomes
\[
\mathcal{L}_{\text{reg}}(\theta) \;=\; \mathcal{L}_{\text{train}}(\theta) \;+\; \lambda \|\theta\|_2^2,
\]
where $\lambda > 0$ controls how strongly large weights are discouraged. This single line connects two bridge topics: the loss from this chapter and the norm from Bridge~I.

\subsection*{Projection Intuition For Least Squares}

Least squares can be read as a projection. No predictor in the allowed linear class may match the target outputs perfectly. Training finds the closest allowed predictor under squared error.

That is a deeper way to read regression. The model is not recovering every detail of the world. It is finding the best approximation available inside the chosen family.

The vocabulary that survives this section: predictor as function, risk as a rule on predictors, norm as size or complexity measure, projection as fitting intuition---and convexity of the risk functional as the formal license for ensembling and model averaging.

\section*{Gradients And Optimization}

Once the training objective is defined, optimization asks how to reduce it. A derivative tells how one quantity changes when another changes slightly. A gradient collects those directional sensitivities across many parameters.

If the loss is \(\mathcal{L}(\theta)\), the gradient
\[
\nabla_\theta \mathcal{L}
\]
points in the direction of increasing loss. To reduce the loss, gradient descent moves the parameters in the opposite direction:
\[
\theta \leftarrow \theta - \eta \nabla_\theta \mathcal{L},
\]
where \(\eta\) is the learning rate.

The mental model is simple:

\begin{itemize}[leftmargin=*]
\item change a parameter slightly,
\item see how the loss responds,
\item collect those responses,
\item move against the direction that increases loss.
\end{itemize}

\begin{codeexample}
theta = initial_parameters()
for step in range(num_steps):
    gradient = grad(training_loss, theta)
    theta = theta - learning_rate * gradient
\end{codeexample}

The pseudocode is short because the mechanism is short.

In practice, gradients are almost never computed on the full dataset. The training set is split into small \emph{mini-batches}, and each update uses the gradient estimated from one batch. This variant, \emph{stochastic gradient descent} (SGD), is noisier per step but far cheaper, and the noise can even help escape poor local minima. The principle---step opposite to the gradient---stays the same.

\begin{codeexample}[caption={From scratch: finite-difference gradient check}]
def finite_difference_gradient(loss_fn, theta, h=1e-5):
    """Numerical gradient of loss_fn at theta, one parameter at a time."""
    grad = [0.0] * len(theta)
    for i in range(len(theta)):
        plus  = list(theta); plus[i]  += h
        minus = list(theta); minus[i] -= h
        grad[i] = (loss_fn(plus) - loss_fn(minus)) / (2 * h)
    return grad
\end{codeexample}

This is the single most useful tool for verifying that a hand-derived gradient is correct. Compute the analytic gradient on a tiny problem; compute the finite-difference gradient with the function above; compare. They should match to four or five significant figures for well-conditioned losses. If they do not, the analytic derivation is wrong, the loss has a bug, or the chosen $h$ is so small that floating-point rounding dominates the difference. Most ``training does not converge'' issues in a from-scratch implementation are silent gradient bugs that this five-line check would have caught.

\begin{example}[One Gradient Descent Step by Hand]
Let the loss function be \(L(w) = (w - 3)^2\), a parabola with minimum at \(w = 3\). Start at \(w = 1\) with learning rate \(\eta = 0.4\).

\textbf{Step 1: Compute the gradient.}
\[
\frac{dL}{dw} = 2(w - 3) = 2(1 - 3) = -4.
\]

\textbf{Step 2: Apply the update rule.}
\[
w_{\text{new}} = w - \eta \cdot \frac{dL}{dw} = 1 - 0.4 \times (-4) = 1 + 1.6 = 2.6.
\]

The weight moved from \(1\) toward \(3\), the minimum. At \(w=1\), the gradient points uphill in the negative direction; subtracting it moves \(w\) downhill toward the minimum, closing the gap.

\begin{center}
\begin{tikzpicture}[font=\small, x=1.6cm, y=1.3cm]
  % Axes
  \draw[->] (-0.2, 0) -- (5.8, 0) node[right] {$w$};
  \draw[->] (0, -0.2) -- (0, 4.5) node[above] {$L(w)$};
  % Parabola: L(w) = (w-3)^2, sampled from w=0.5 to w=5.5
  \draw[thick] plot[domain=0.5:5.5, samples=80] (\x, {(\x-3)*(\x-3)});
  % x-axis tick: start
  \draw (1, 0.05) -- (1, -0.05);
  \node[below, font=\footnotesize] at (1, -0.1) {$w{=}1$};
  % x-axis tick: minimum
  \draw (3, 0.05) -- (3, -0.05);
  \node[below, font=\footnotesize] at (3, -0.1) {$w{=}3$};
  % Points on the parabola
  \filldraw[fill=white, draw=black] (1, 4) circle (2pt)
      node[above right] {$(1,\,4)$};
  \filldraw[fill=black] (2.6, 0.16) circle (2pt)
      node[right, xshift=4pt, font=\footnotesize] {$(2.6,\,0.16)$};
  \filldraw[fill=black] (3, 0) circle (2pt);
  % Arrow showing the step along the x-axis
  \draw[->, thick] (1, -0.6) -- (2.55, -0.6)
        node[midway, below, font=\footnotesize] {step $= 1.6$};
\end{tikzpicture}
\end{center}

The open circle marks the starting point \(w=1\); the filled circles mark the updated weight \(w=2.6\) and the minimum \(w=3\). One gradient step more than halved the distance to the optimum. A smaller learning rate would shorten the step; a larger one might overshoot.
\end{example}

Most of the practical difficulty in training comes from scale, instability, architecture choice, and noisy data---not from the basic update rule itself.

\section*{Chain Rule As Credit Assignment}

The chain rule matters because many models are compositions of simpler computations. If
\[
u = a z + b,
\qquad
z = c x,
\qquad
\mathcal{L} = (u-y)^2,
\]
the loss depends on \(c\) indirectly through \(z\) and \(u\). The chain rule lets the model pass responsibility backward through the computation.

The derivative with respect to the early parameter \(c\) is
\[
\frac{\partial \mathcal{L}}{\partial c}
=
\frac{\partial \mathcal{L}}{\partial u}
\cdot \frac{\partial u}{\partial z}
\cdot \frac{\partial z}{\partial c}
=
2(u-y)\cdot a \cdot x.
\]
So if \(a=2\), \(x=3\), and \(u-y=1\), the gradient is \(12\): a small change in \(c\) has a clear local effect on the loss.

The useful reading:

\begin{quote}
how much did an earlier quantity contribute to the final error, and how much credit or blame should move backward?
\end{quote}

This is why backpropagation is not calculus drill. It is structured credit assignment through a composed function.

\section*{Generative vs.\ Discriminative: Two Ways to Use Probability}

All the models discussed so far learn to predict a label from an input: $P(Y \mid X)$. This is the \emph{discriminative} approach---the model discriminates between classes based on the observed features. Logistic regression, neural networks, and decision trees are all discriminative.

A \emph{generative} model takes a fundamentally different approach. It learns how the data is produced: $P(X \mid Y)$ (what does a typical input look like, given the class?) and $P(Y)$ (how common is each class?). To classify a new input, Bayes' theorem combines these:
\[
P(Y \mid X) = \frac{P(X \mid Y) \, P(Y)}{P(X)}
\]

The numerator is easy: multiply the class-conditional likelihood by the prior. The denominator $P(X)$ is the same for all classes, so for classification we can ignore it---just pick the class that maximizes $P(X \mid Y) \, P(Y)$.

\textbf{Why does this distinction matter?} It changes what the model learns and what it can do:

\begin{itemize}[leftmargin=*]
\item A discriminative model focuses all its capacity on the decision boundary. It is often more accurate for classification, but it does not model the data distribution, so it has no built-in density for outlier detection or sampling new data.
\item A generative model learns the structure of the data itself. It can detect inputs unlike anything it has seen (outlier detection), handle missing features by marginalizing over them, and generate new examples. The cost is stronger assumptions about the data distribution.
\end{itemize}

\begin{example}[Generative Thinking for Spam]
A discriminative spam filter asks: given these word frequencies, is this message spam? A generative spam filter asks: what do spam messages look like, and what do legitimate messages look like? The generative model can flag an email written in a language it has never seen---it is unlikely under \emph{both} models, so it looks out-of-distribution. A plain closed-set discriminative classifier has no built-in mechanism for this; it needs an added rejection rule, uncertainty check, or out-of-distribution detector.
\end{example}

\textbf{Priors, posteriors, and MAP.} In Bayesian vocabulary, $P(Y)$ is the \emph{prior}---your belief about class probabilities before seeing the data. $P(Y \mid X)$ is the \emph{posterior}---your updated belief after seeing the input. The same logic applies to model parameters: a \emph{prior} on parameters $P(\theta)$ expresses which parameter values you believe are plausible before training. \emph{Maximum a posteriori} (MAP) estimation finds the parameters maximizing $P(\theta \mid \text{data}) \propto P(\text{data} \mid \theta) \, P(\theta)$. This gives a principled interpretation of many regularizers: in penalized likelihood, an $L_2$ penalty corresponds to a Gaussian prior, and an $L_1$ penalty to a Laplace prior. What looked like an ad hoc penalty reads as a prior belief when the modeling assumptions match.

These ideas---generative models, priors, posteriors, MAP---return in full force in Chapter~\ref{ch:probabilistic-models} (probabilistic models), Chapter~\ref{ch:structure-beyond-linearity} (Naive Bayes), and Chapter~\ref{ch:generative-representations} (generative representations). This bridge gives just enough vocabulary to make those chapters readable.

\textbf{Discriminative or generative?} When designing a system, ask: do I only need to classify, or do I also need to detect the unfamiliar, handle missing information, or generate new examples? The answer decides whether a discriminative model suffices or whether a generative perspective is needed.

\section*{Chapter Artifact: Learning-Signal Card}

\begin{quote}
\noindent\textbf{Learning-Signal Card}

\medskip
\begin{tabular}{p{2.6cm}p{5.4cm}}
\toprule
Field & What to write \\
\midrule
Model output & Score, probability, ranking signal, class label, or numeric prediction \\
Pointwise loss & The penalty on one example and what behavior it rewards \\
Dataset objective & The average or aggregate quantity minimized during training \\
Reported metric & The number used to justify the final claim \\
Decision rule & Threshold, ranking cutoff, abstention rule, or review policy \\
Main mismatch risk & What the loss may optimize that the deployment goal does not actually want \\
\bottomrule
\end{tabular}
\end{quote}

This card should make it hard to say ``the model improved'' without saying what improved, under which loss, and in service of which decision.

\section*{Common Mistakes}

\begin{mistakebox}
\item \textbf{Treating a score as a calibrated probability.} Check what the number means before interpreting it as a probability.
\item \textbf{Treating lower training loss as deployment success.} The training objective can improve while the real task stays the same or gets worse.
\item \textbf{Choosing a loss without asking what wrong behavior it permits.} Loss choice is part of task design, not a cosmetic detail.
\item \textbf{Forgetting that a model is a function used on inputs.} The predictor lives on examples, not only as a parameter vector stored in memory.
\item \textbf{Treating gradients as mysterious symbols instead of local direction information.} Gradients say how to move parameters to reduce loss.
\end{mistakebox}

\section*{Looking Ahead}

The two bridge chapters were not meant as isolated mathematics review. Their job was to make later modeling language readable: representation in the first bridge; score, probability, decision rule, loss, and update signal in this one. Chapter 1 now begins the main argument of the book by asking what decision, proxy target, data source, and held-out evidence make a learning problem worth modeling at all.

\begin{example}[Joint Artifact: Representation and Learning Signal Together]
For a spam-filtering task, the two bridge artifacts work together. The \emph{representation card} (Bridge~I) names the features: word frequencies, sender score, and link count, each stored as a coordinate in $\mathbb{R}^{50}$. The \emph{learning-signal card} (Bridge~II) specifies the loss (log loss), the decision threshold ($\tau = 0.3$ because false negatives are costly), and the baseline (predict the majority class). The two cards define what the model sees and how it learns---before any algorithm is chosen.
\end{example}

\begin{chaptersummary}
\item A score, a probability, a decision rule, a loss, and a dataset-level objective are different objects with different jobs.
\item Conditional probability expresses uncertainty given observed evidence.
\item Expectation is the language of average consequence.
\item Logarithms and exponentials appear because they make probability-based modeling and optimization easier to express.
\item A predictor is a function, and empirical risk is a rule that assigns one number to the whole predictor.
\item Gradients and the chain rule turn a loss into a usable update signal.
\item The chapter's main deliverable is a learning-signal card, written before claiming that a training objective and a deployment goal are aligned.
\item Generative models learn $P(X \mid Y)$ and can detect outliers and generate data; discriminative models learn $P(Y \mid X)$ and focus on the decision boundary.
\item MAP estimation connects regularization to Bayesian prior beliefs about parameter values.
\item Probability is already part of the model: choosing a likelihood, a loss, or a decision threshold is a modeling commitment about what counts as evidence, not a downstream technicality.
\end{chaptersummary}

\begin{table}[h]
\centering
\small
\caption{Where probability and optimization ideas return in later chapters.}
\begin{tabular}{@{}L{3.05cm}L{6.75cm}@{}}
\toprule
\textbf{Bridge Topic} & \textbf{Used In} \\
\midrule
Conditional probability and Bayes' rule & Chapters~\ref{ch:framing-learning-problems},~\ref{ch:language-grounding} (framing, language grounding) \\
Loss functions and scoring rules & Chapters~\ref{ch:prediction-loss-decision},~\ref{ch:optimization-neural-networks} (metrics, neural training) \\
Empirical vs.\ population risk & Chapters~\ref{ch:evaluation-discipline},~\ref{ch:experiments-claims} (evaluation, experiments) \\
Gradient descent & Chapter~\ref{ch:optimization-neural-networks} (optimization, backpropagation) \\
Cost-sensitive thresholds & Chapters~\ref{ch:prediction-loss-decision},~\ref{ch:vision} (decision rules, detection) \\
Generative vs.\ discriminative, priors, MAP & Chapters~\ref{ch:structure-beyond-linearity},~\ref{ch:probabilistic-models},~\ref{ch:generative-representations} (Naive Bayes, probabilistic models, generative representations) \\
Expectation under a distribution & Chapters~\ref{ch:probabilistic-models},~\ref{ch:reinforcementlearning},~\ref{ch:generative-representations} (variational objectives, policy gradients, ELBO) \\
Jensen's inequality & Chapters~\ref{ch:probabilistic-models},~\ref{ch:generative-representations} (EM monotonicity, ELBO derivation) \\
Multivariate Gaussian density & Chapters~\ref{ch:probabilistic-models},~\ref{ch:generative-representations} (GMM, Bayesian linear regression, VAE prior, diffusion noise) \\
\bottomrule
\end{tabular}
\end{table}

\begin{studyquestions}
\item Why is a model score not automatically the same thing as a probability?
\item What is the practical difference between a pointwise loss and empirical risk?
\item In plain language, what is the functional-analysis viewpoint on a loss function that this book needs?
\item Why is the chain rule better understood as credit assignment than as symbol pushing?
\item What can a generative model do that a discriminative model cannot, and why does this come at a cost?
\end{studyquestions}

\begin{exercises}
\item \exercisetag{Concept}{Core}Give one example of a score that should be used for ranking but not treated as a calibrated probability.
\item \exercisetag{Calculation}{Core}Compute the expected value of an action that gives \(+4\) utility when a positive case is correctly caught and \(-2\) utility when a false alarm occurs, if the predicted probability of a true positive case is \(0.7\).
\item \exercisetag{Calculation}{Core}Write down one pointwise squared-error loss and one pointwise log-loss expression, then explain in words what each one punishes.
\item \exercisetag{Concept}{Core}Explain the difference between ``choose parameters'' and ``choose a predictor from a hypothesis class'' in one paragraph.
\item \exercisetag{Concept}{Intermediate}Fill out a learning-signal card for spam filtering, student support, or binary image classification. If you plan to carry one running case through the book, keep the same case and add one sentence describing what downstream decision or review action the score should eventually support.
\item \exercisetag{Implementation}{Intermediate}One SGD step on log loss by hand. For a logistic model on two features with weights $w = (0.0, 0.0)$, bias $b = 0$, and one training example $x = (1.0, 2.0)$ with label $y = 1$, compute (a) the predicted probability $p = \sigma(w^\top x + b)$, (b) the per-parameter gradient of the log loss $-\log p$ at this point (use the identity $\partial \log p / \partial w_j = (p - y)\, x_j$), and (c) the updated weights and bias after one SGD step with learning rate $\eta = 0.1$. Then verify your analytic gradient against the finite-difference gradient check.
\end{exercises}

\begin{challengeexercises}
\item \exercisetag{Design}{Advanced}Construct a tiny example in which two models have similar accuracy but very different log loss, and explain what that says about confidence.
\item \exercisetag{Concept}{Advanced}Give a plain-language explanation of least squares as projection into an allowed function class.
\item \exercisetag{Design}{Advanced}Design a toy two-stage computation and describe how the chain rule would pass credit backward through it without computing every derivative explicitly.
\end{challengeexercises}
