\chapter*{Part Synthesis: Evidence, Reliability, and Systems}
\addcontentsline{toc}{chapter}{Part Synthesis: Evidence, Reliability, and Systems}

This part turned mature ML practice into explicit undergraduate workflow. Students now have one chain from experimental claim to reliability review to launch-or-no-launch service judgment, which is exactly where the book's problem-driven identity becomes most distinctive.

\begin{chaptersummary}
\item \textbf{Question this part taught.} What does the result justify, where can it still fail badly enough to matter, and what must be true before a model becomes a real system?
\item \textbf{Artifact chain this part added.} Experiment claim sheet $\rightarrow$ reliability review card $\rightarrow$ security threat card $\rightarrow$ system readiness brief.
\item \textbf{What a student can now do.} Turn an offline result into a bounded scientific claim, test that claim against reliability and adversarial hazards, and make a technical launch or no-launch recommendation for the whole service.
\end{chaptersummary}

Both running cases---the course-support assistant and the pedestrian detector---illustrate the same arc: a claim must be bounded before it is believed (Chapter~\ref{ch:experiments-claims}), reliability must be audited across subgroups, shift conditions, uncertainty signals, explanation limits, and adversarial pressure before the system is trusted (Chapter~\ref{ch:reliability}), and the full system must meet operational requirements before it is launched (Chapter~\ref{ch:models-systems}). The artifacts from this part (claim sheet, reliability review card, security threat card, and system readiness brief) are designed to be carried forward into professional practice, not just submitted as coursework.
