\chapter*{Part Synthesis: Modalities}
\addcontentsline{toc}{chapter}{Part Synthesis: Modalities}

\begin{whychaptermatters}
This part made the book's general workflow modality-specific. Vision sharpened invariance, language sharpened context and grounding, and sequential modeling sharpened prediction-time legality. The common lesson is that different data types still demand explicit audits before architecture choice can be trusted.
\end{whychaptermatters}

\begin{chaptersummary}
\item \textbf{Question this part taught.} What task-specific structure must be preserved, what nuisance variation should be ignored, and what evidence makes a modality-specific claim honest under deployment conditions?
\item \textbf{Artifact chain this part added.} Vision design and stress-test card $\rightarrow$ NLP context-and-grounding card $\rightarrow$ sequential design card.
\item \textbf{What a student can now do.} Match invariance, context, grounding, and prediction-time audits to the modality instead of treating pixels, tokens, and sequences as interchangeable inputs.
\end{chaptersummary}

Three shared challenges recur across all modalities. First, benchmark success does not guarantee deployment success: shortcut learning in vision, fluent-but-false generation in language, and temporal leakage in sequences all produce misleading offline numbers. Second, representation choices are task-specific, not universal. Third, evaluation must respect deployment constraints---camera conditions, grounding requirements, and prediction-time boundaries. The running cases (pedestrian detector, course-support assistant, wake-word detector, and demand forecaster) illustrate that the same disciplined audit applies regardless of data type.

