\chapter*{Part Synthesis: Deep Learning}
\addcontentsline{toc}{chapter}{Part Synthesis: Deep Learning}

\begin{whychaptermatters}
This part widened the modeling story from engineered structure to learned and reusable structure. Students should now be able to debug a small neural workflow, separate optimization from generalization trouble, and attribute improvements to optimization, regularization, or representation quality. The course-support assistant now serves as a concrete case for representation reuse: can a general language representation, pretrained on broad text, be adapted to handle student queries about deadlines, policies, and course logistics? Part~III applies these ideas to specific data modalities---vision, language, and sequences---where representation choices become even more consequential.
\end{whychaptermatters}

\begin{chaptersummary}
\item \textbf{Question this part taught.} When should structure be learned rather than engineered, when is a learned representation reusable enough to support later tasks honestly, and when must a system learn from interaction rather than static prediction?
\item \textbf{Artifact chain this part added.} Minimum credible neural training report $\rightarrow$ reuse and adaptation plan $\rightarrow$ interaction design card.
\item \textbf{What a student can now do.} Train and diagnose a small network, make a cleaner transfer claim by separating the reusable space, new evidence, and downstream adaptation depth, and design an RL system with an audited reward signal and documented exploration strategy.
\end{chaptersummary}

