\chapter*{Part Synthesis: Models and Representations}
\addcontentsline{toc}{chapter}{Part Synthesis: Models and Representations}

\begin{whychaptermatters}
This part widened the modeling story from engineered structure to learned, probabilistic, reusable, generative, and interactive structure. Students should now be able to debug a small neural workflow, ask what uncertainty the model can report, separate representation reuse from downstream adaptation, evaluate generative claims cautiously, and recognize when feedback from actions changes the learning problem.
\end{whychaptermatters}

\begin{chaptersummary}
\item \textbf{Question this part taught.} When should structure be learned rather than engineered, what uncertainty does the model know how to express, when is a learned representation reusable, what can a generative model honestly claim, and when must a system learn from interaction rather than static prediction?
\item \textbf{Artifact chain this part added.} Minimum credible neural training report $\rightarrow$ uncertainty audit card $\rightarrow$ reuse and adaptation plan $\rightarrow$ generative model audit card $\rightarrow$ interaction design card.
\item \textbf{What a student can now do.} Train and diagnose a small network, distinguish optimization from uncertainty and representation quality, make a cleaner transfer claim, audit generated data before using it as evidence, and design an RL system with a documented reward signal and exploration strategy.
\end{chaptersummary}

The course-support assistant now serves as a concrete case for representation reuse: can a general language representation, pretrained on broad text, be adapted to handle student queries about deadlines, policies, and course logistics? The pedestrian detector supplies the complementary case for uncertainty and generation: can the model recognize when the visual evidence is unfamiliar, and can generated or augmented data improve performance without hiding slice failures? Part~III applies these ideas to specific data modalities---vision, language, sequences, and recommendation---where representation choices become operational choices about what the system sees, ranks, exposes, and acts on.
