\chapter*{Part Synthesis: Foundations}
\addcontentsline{toc}{chapter}{Part Synthesis: Foundations}

This part built the book's first complete workflow. Students now have a way to frame a decision, choose what good behavior means, protect the claim with held-out evidence, test a serious baseline, diagnose what structure is still missing, and design the dataset that makes the learning problem possible.

\begin{chaptersummary}
\item \textbf{Question this part taught.} What problem are we solving, what counts as good behavior, what evidence protects the claim, what missing structure justifies the next model family, and what data design makes the task learnable?
\item \textbf{Artifact chain this part added.} Framing memo $\rightarrow$ metric-and-action worksheet $\rightarrow$ evaluation plan $\rightarrow$ first baseline design sheet $\rightarrow$ structure diagnosis sheet $\rightarrow$ data design card.
\item \textbf{What a student can now do.} Turn a vague ML request into a defended baseline workflow, justify the next model family from failure type instead of from fashion, and document how data collection, labeling, and label-acquisition strategy shape what the model can learn.
\end{chaptersummary}

The course-support assistant has been the primary running case throughout Part~I, grounding every artifact in a concrete decision problem. The pedestrian detector has also appeared as a reminder that the same workflow must survive outside tabular or text-heavy cases. Part~II asks a harder question: when hand-engineered features and classical structure reach their limits, can a model learn useful internal representations while still reporting honest uncertainty? The answer takes us through neural networks (Chapter~\ref{ch:optimization-neural-networks}), probabilistic models (Chapter~\ref{ch:probabilistic-models}), representation reuse (Chapter~\ref{ch:representation-foundation-models}), and generative representations (Chapter~\ref{ch:generative-representations}). Part~III then steps outside the prediction-from-data thread to learning from interaction (Chapter~\ref{ch:reinforcementlearning}), where the agent's own actions shape the data it later sees.
